---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 100
)
```

# eigencore

**eigencore** computes the top-*k* singular triplets or eigenpairs of a large
sparse or structured matrix in R — the computation behind PCA on big sparse
data, spectral embeddings, LSA, and low-rank approximation.

eigencore focuses on two things:

1. **Every result is checked.** Each call returns residuals, a backward-error
   bound, orthogonality loss, and a single `passed`/`failed` flag. When a bound
   can only be estimated, the certificate says so instead of passing.
2. **Centering and scaling without densifying.** Explicitly centering a sparse
   matrix for PCA can require a dense copy. eigencore solves the centered (or
   scaled, or composed) problem as an operator, without forming that copy.

Supported structured problems run through fast native kernels. The
[Benchmarks](#benchmarks) section provides reproducible single-machine timings
without treating them as cross-package rankings.

## Installation

``` r
# install.packages("pak")
pak::pak("bbuchsbaum/eigencore")
```

## Quick start

The top 10 singular triplets of a 100,000 × 500 sparse matrix — the core
computation in sparse PCA and LSA:

```{r quick-start}
library(eigencore)
library(Matrix)

set.seed(2)
A <- as(rsparsematrix(100000, 500, density = 0.002), "dgCMatrix")

fit <- svd_partial(A, rank = 10, target = largest())
fit
```

The printout names the kernel that ran, gives the worst residual, backward
error, and orthogonality loss across the returned triplets, and shows the
certificate passed with an exact norm bound. This problem uses the native
certified Gram path; see [Benchmarks](#benchmarks) for a reproducible timing on
the development machine.

```{r scree, echo = FALSE, fig.alt = "The ten largest singular values highlighted in blue against the full 500-point singular spectrum of A in grey."}
sig_all <- sort(sqrt(pmax(0, eigen(as.matrix(crossprod(A)),
                                   symmetric = TRUE, only.values = TRUE)$values)),
                decreasing = TRUE)
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(sig_all), sig_all, pch = 19, cex = 0.4, col = "grey78",
     bty = "n", xlab = "rank (largest to smallest)", ylab = "singular value")
points(seq_len(10), sort(fit$d, decreasing = TRUE),
       pch = 19, cex = 1.3, col = "#2166ac")
legend("topright", c("full spectrum (dense reference)", "computed by eigencore"),
       pch = 19, pt.cex = c(0.7, 1.3), col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Certificates

An iterative solver can stop early, miss a cluster, or lose orthogonality and
still return plausible-looking numbers. eigencore makes validation part of the
returned result: it checks both singular relations (`||A v - sigma u||` and
`||A^T u - sigma v||`) and exposes the evidence:

```{r certificate}
fit$certificate
```

When the check cannot be made exact, the certificate says so. For a
**column-centered** sparse matrix the only cheap norm bound is a stochastic
estimate, so eigencore returns the singular values but sets `passed = FALSE`
and tells you why:

```{r withheld}
cen <- svd_partial(center(A, columns = TRUE), rank = 5, target = largest())

cen$certificate$passed
cen$certificate$norm_bound_type
cen$certificate$notes
```

You decide whether an estimated bound is good enough for your analysis. The
explicit flag lets downstream code distinguish an exact certificate from an
estimate without inferring that distinction from solver convergence alone.

## Center and scale without densifying

A dense centered copy of `A` would occupy **400 MB**; the sparse original is a
few MB. `center()` gives you the centered map as an *operator*, and the solver
works through it directly:

```{r operators}
A_centered <- center(A, columns = TRUE)        # a 100000 x 500 operator, not a matrix
svd_partial(A_centered, rank = 5, target = largest())$d
```

Build operators with `linear_operator()`, combine them with `compose()`,
`crossprod_operator()`, `scale_cols()`, `center()`, and friends. The planner
picks the kernel from the structure; `plan_solver()` shows the choice before
you commit to a long solve:

```{r plan}
plan_solver(svd_problem(A_centered, target = largest()), rank = 5)$method
```

## Smallest eigenvalues of a symmetric operator

The same interface handles symmetric eigenproblems. Here is a sparse
second-difference operator (a 1-D graph Laplacian) of size 20,000, asking for
its **smallest** eigenvalues — the hard end of the spectrum for iterative
solvers:

```{r eigen}
n <- 20000
L <- bandSparse(n, n, k = c(-1, 0, 1),
                diagonals = list(rep(-1, n - 1), rep(2, n), rep(-1, n - 1)))
L <- as(L, "dgCMatrix")

eig <- eig_partial(L, k = 8, target = smallest())
eig
```

The planner selects a native tridiagonal shift-invert path and the resulting
certificate passes. The exact spectrum is known in closed form, so the answer
can also be checked directly:

```{r spectrum, echo = FALSE, fig.alt = "The eight smallest eigenvalues highlighted in blue at the bottom of the full analytic spectrum of the 20,000-point 1-D Laplacian shown in grey."}
lambda_all <- 4 * sin(seq_len(n) * pi / (2 * (n + 1)))^2  # exact Laplacian spectrum
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(lambda_all), lambda_all, type = "l", col = "grey78", lwd = 1.4,
     bty = "n", xlab = "rank (smallest to largest)", ylab = "eigenvalue")
points(seq_along(eig$values), sort(eig$values), pch = 19, cex = 1.2, col = "#2166ac")
legend("topleft", c("full spectrum (analytic)", "computed by eigencore"),
       lwd = c(1.4, NA), pch = c(NA, 19), pt.cex = 1.2,
       col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Benchmarks {#benchmarks}

These are median wall-clock times for eigencore on one development machine.
They include certificate computation and are intended as a reproducible
performance smoke test, not a cross-package ranking or a performance guarantee.
Reproduce them with `Rscript inst/benchmarks/bench-readme.R`.

| Problem (certificate `passed`) | Median time | Planner path |
|---|---:|---|
| Tall sparse SVD, 100000 × 500, k = 10 | 15 ms | native certified Gram SVD special case |
| Wide sparse SVD, 500 × 100000, k = 10 | 12 ms | native certified Gram SVD special case |
| Banded Hermitian, smallest, n = 20000, k = 8 | 31 ms | native tridiagonal Hermitian shift-invert |

<sub>Measured with R 4.5.1 on aarch64-apple-darwin20. Timings depend on the
processor, BLAS/LAPACK, package versions, sparsity pattern, and workload; rerun
the script before making performance decisions.</sub>

The SVD rows use a bounded Gram kernel for tall or wide sparse problems whose
small dimension is ≤ 512 (≤ 1024 for wide matrices). Other shapes may select
different paths with different costs. `fit$method` always names the path that
ran, making the relevant implementation boundary visible.

That path has fixed work associated with solving the smaller Gram problem and
constructing a certified result. On the tiny examples used in documentation
smoke tests, this overhead can dominate and `RSpectra` or `irlba` may finish
sooner. As the long dimension grows while the smaller dimension remains within
the planner boundary, the same path can cross over and become faster. This is a
shape-specific effect, not a claim that eigencore becomes faster whenever a
matrix gets larger; the [benchmark vignette](https://bbuchsbaum.github.io/eigencore/articles/benchmarks.html)
explains the comparison and its limits.

## When to use what

Use **eigencore** for tall or wide sparse SVD (PCA-shaped problems), the
smallest eigenvalues of banded or structured symmetric operators (certified,
with automatic planner selection), centered or scaled or composed operators,
dense generalized eigen/QZ/GSVD compatibility work through `eig_full()`,
`generalized_schur()`, and `generalized_svd()`, and workflows where explicit
certificate metadata is useful.

For workloads outside those structured paths, benchmark the candidate packages
on your own matrices and hardware rather than assuming any implementation will
be fastest. Side-by-side evaluation is straightforward because eigencore ships
RSpectra-compatible wrappers with the same arguments:

```{r rspectra-shim}
res <- eigs_sym(L, k = 8, which = "SA")
res$values
res$certificate$passed
```

`eigs()`, `eigs_sym()`, and `svds()` accept the same `which` codes as RSpectra
(`"LM"`, `"SM"`, `"LA"`, `"SA"`, `"LR"`, `"SR"`, `"LI"`, `"SI"`, `"BE"`) and
additionally return a certificate.

## Learning more

`vignette("eigencore", package = "eigencore")` is the guided tour.
`vignette("certificates", package = "eigencore")` explains how to read the
numerical evidence and what to do when a check fails.

## Status

eigencore 1.0.0 is headed for CRAN. The exported API is stable — it is frozen
by a snapshot test, and breaking changes follow semantic versioning from here.
Numerical behavior is covered by adversarial and reference tests, while solver
results expose certificate and planner provenance on every call. Problem
classes without a native kernel yet — general matrix-free SVD, interior
eigenvalues at scale, nonsymmetric Krylov–Schur — are labeled `reference` in
`fit$method`. The package vignettes describe the workflow map, benchmark
evidence, and current boundaries.

## License

MIT © Bradley Buchsbaum.
