---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 100
)
```

# eigencore

**eigencore** computes the top-*k* singular triplets or eigenpairs of a large
sparse or structured matrix in R — the computation behind PCA on big sparse
data, spectral embeddings, LSA, and low-rank approximation. It runs on native
C++ kernels and attaches a numerical certificate to every result, so you know
the answer is right, not just that the solver stopped.

If you use **RSpectra**, **irlba**, or **PRIMME** for these problems, eigencore
does the same job and adds two things:

1. **Every result is checked.** Each call returns residuals, a backward-error
   bound, orthogonality loss, and a single `passed`/`failed` flag. When a bound
   can only be estimated, the certificate says so instead of passing.
2. **Centering and scaling without densifying.** Centering a sparse matrix for
   PCA normally forces a dense copy. eigencore solves the centered (or scaled,
   or composed) problem directly — the dense matrix is never formed.

On PCA-shaped sparse SVD problems it is also 1.6×–24× faster than RSpectra,
irlba, and PRIMME (see [Benchmarks](#benchmarks)).

## Installation

``` r
# install.packages("pak")
pak::pak("bbuchsbaum/eigencore")
```

## Quick start

The top 10 singular triplets of a 100,000 × 500 sparse matrix — the core
computation in sparse PCA and LSA:

```{r quick-start}
library(eigencore)
library(Matrix)

set.seed(2)
A <- as(rsparsematrix(100000, 500, density = 0.002), "dgCMatrix")

fit <- svd_partial(A, rank = 10, target = largest())
fit
```

The printout names the kernel that ran, gives the worst residual, backward
error, and orthogonality loss across the returned triplets, and shows the
certificate passed with an exact norm bound. The solve takes 15 ms — 1.6×
faster than RSpectra, 9× faster than irlba, 2.7× faster than PRIMME
(see [Benchmarks](#benchmarks)).

```{r scree, echo = FALSE, fig.alt = "The ten largest singular values highlighted in blue against the full 500-point singular spectrum of A in grey."}
sig_all <- sort(sqrt(pmax(0, eigen(as.matrix(crossprod(A)),
                                   symmetric = TRUE, only.values = TRUE)$values)),
                decreasing = TRUE)
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(sig_all), sig_all, pch = 19, cex = 0.4, col = "grey78",
     bty = "n", xlab = "rank (largest to smallest)", ylab = "singular value")
points(seq_len(10), sort(fit$d, decreasing = TRUE),
       pch = 19, cex = 1.3, col = "#2166ac")
legend("topright", c("full spectrum (dense reference)", "computed by eigencore"),
       pch = 19, pt.cex = c(0.7, 1.3), col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Certificates

An iterative solver can stop early, miss a cluster, or lose orthogonality and
still return plausible-looking numbers. RSpectra, irlba, and PRIMME hand you
vectors and values; checking them is your problem. eigencore checks both
singular relations (`||A v - sigma u||` and `||A^T u - sigma v||`) and hands
you the evidence:

```{r certificate}
fit$certificate
```

When the check cannot be made exact, the certificate says so. For a
**column-centered** sparse matrix the only cheap norm bound is a stochastic
estimate, so eigencore returns the singular values but sets `passed = FALSE`
and tells you why:

```{r withheld}
cen <- svd_partial(center(A, columns = TRUE), rank = 5, target = largest())

cen$certificate$passed
cen$certificate$norm_bound_type
cen$certificate$notes
```

You decide whether an estimated bound is good enough for your analysis. Other
solvers don't give you the choice — they return the same numbers with no flag.

## Center and scale without densifying

A dense centered copy of `A` would occupy **400 MB**; the sparse original is a
few MB. `center()` gives you the centered map as an *operator*, and the solver
works through it directly:

```{r operators}
A_centered <- center(A, columns = TRUE)        # a 100000 x 500 operator, not a matrix
svd_partial(A_centered, rank = 5, target = largest())$d
```

Build operators with `linear_operator()`, combine them with `compose()`,
`crossprod_operator()`, `scale_cols()`, `center()`, and friends. The planner
picks the kernel from the structure; `plan_solver()` shows the choice before
you commit to a long solve:

```{r plan}
plan_solver(svd_problem(A_centered, target = largest()), rank = 5)$method
```

## Smallest eigenvalues of a symmetric operator

The same interface handles symmetric eigenproblems. Here is a sparse
second-difference operator (a 1-D graph Laplacian) of size 20,000, asking for
its **smallest** eigenvalues — the hard end of the spectrum for iterative
solvers:

```{r eigen}
n <- 20000
L <- bandSparse(n, n, k = c(-1, 0, 1),
                diagonals = list(rep(-1, n - 1), rep(2, n), rep(-1, n - 1)))
L <- as(L, "dgCMatrix")

eig <- eig_partial(L, k = 8, target = smallest())
eig
```

eigencore solves this in 31 ms, certificate passed. The same call through
RSpectra's default mode (`which = "SA"`) does not converge on this matrix —
0 of 8 eigenvalues after 2.7 s. RSpectra is fast here if you know to switch
it to shift-invert mode (`sigma = 0` solves it in 13 ms); eigencore makes
that choice for you and certifies the result. The exact spectrum is known in
closed form, so the answer can be checked directly:

```{r spectrum, echo = FALSE, fig.alt = "The eight smallest eigenvalues highlighted in blue at the bottom of the full analytic spectrum of the 20,000-point 1-D Laplacian shown in grey."}
lambda_all <- 4 * sin(seq_len(n) * pi / (2 * (n + 1)))^2  # exact Laplacian spectrum
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(lambda_all), lambda_all, type = "l", col = "grey78", lwd = 1.4,
     bty = "n", xlab = "rank (smallest to largest)", ylab = "eigenvalue")
points(seq_along(eig$values), sort(eig$values), pch = 19, cex = 1.2, col = "#2166ac")
legend("topleft", c("full spectrum (analytic)", "computed by eigencore"),
       lwd = c(1.4, NA), pch = c(NA, 19), pt.cex = 1.2,
       col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Benchmarks {#benchmarks}

Median wall-clock time. eigencore's column includes computing the certificate;
the baseline columns are solve only. Reproduce with
`Rscript inst/benchmarks/bench-readme.R`.

| Problem (all certified `passed` by eigencore) | eigencore | vs RSpectra | vs irlba | vs PRIMME |
|---|---|---|---|---|
| Tall sparse SVD, 100000 × 500, k = 10 | 15 ms | **1.6× faster** | **9.1× faster** | **2.7× faster** |
| Wide sparse SVD, 500 × 100000, k = 10 | 12 ms | **12.0× faster** | **24.1× faster** | **14.0× faster** |
| Banded Hermitian, smallest, n = 20000, k = 8 | 31 ms | see note¹ | — | — |

<sub>R 4.5.1 · aarch64-apple-darwin20 · RSpectra 0.16.2 / irlba 2.3.7 /
PRIMME 3.2.6. The SVD rows also allocate ~6–8× less memory than irlba. Rerun
the script for numbers on your machine and BLAS.</sub>

<sub>¹ RSpectra's default mode (`which = "SA"`) returns 0 of 8 eigenvalues on
this matrix after 2.7 s. Its shift-invert mode (`sigma = 0`) solves it in
13 ms — faster than eigencore — if you know to reach for it. eigencore picks
the method itself and certifies the result.</sub>

The SVD speedups come from a certified Gram kernel for tall/wide sparse
problems whose small dimension is ≤ 512 (≤ 1024 for wide matrices). Outside
that — a tall SVD with more than 512 columns, or a general sparse matrix like
a 2-D-grid Laplacian — RSpectra is currently 2–6.5× faster on the cases we
measure, narrowing to parity on the larger grid (n = 40,000). `fit$method`
always names the path that ran, so there is no guessing.

## When to use what

Use **eigencore** for tall or wide sparse SVD (PCA-shaped problems), the
smallest eigenvalues of banded or structured symmetric operators (certified,
no mode tuning), centered or scaled or composed operators, dense generalized
eigen/QZ/GSVD compatibility work through `eig_full()`, `generalized_schur()`,
and `generalized_svd()`, and anywhere you want the result checked rather than
taken on faith.

Use **RSpectra or irlba** for general large sparse problems outside those
shapes, and RSpectra's shift-invert mode when raw speed on smallest or
interior eigenvalues matters more than a certificate. Switching in either
direction is a one-line change, because eigencore ships drop-in wrappers with
the same arguments:

```{r rspectra-shim}
res <- eigs_sym(L, k = 8, which = "SA")
res$values
res$certificate$passed
```

`eigs()`, `eigs_sym()`, and `svds()` accept the same `which` codes as RSpectra
(`"LM"`, `"SM"`, `"LA"`, `"SA"`, `"LR"`, `"SR"`, `"LI"`, `"SI"`, `"BE"`) and
additionally return a certificate.

## Learning more

`vignette("eigencore", package = "eigencore")` is the guided tour.
`vignette("certificates", package = "eigencore")` explains how to read the
numerical evidence and what to do when a check fails.

## Status

eigencore 1.0.0 is headed for CRAN. The exported API is stable — it is frozen
by a snapshot test, and breaking changes follow semantic versioning from here.
The numerics are solid: every shipped solver path is benchmarked against
RSpectra and PRIMME, covered by an adversarial test suite, and certified on
every call. Problem classes without a native kernel yet — general matrix-free
SVD, interior eigenvalues at scale, nonsymmetric Krylov–Schur — are labeled
`reference` in `fit$method` rather than quietly slow. The package vignettes
describe the workflow map, benchmark evidence, and current boundaries.

## License

MIT © Bradley Buchsbaum.
