---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 100
)
```

# eigencore

**eigencore** is a native C++ engine for partial eigenvalue and singular-value
computation in R — the top-*k* singular triplets or eigenpairs of a large sparse
or structured matrix, with a numerical certificate attached to each result.

It is **not** a replacement for base `svd()` / `eigen()`, which compute the
*full* dense decomposition of a small matrix. It is an alternative to the partial
solvers you might already use — **RSpectra**, **irlba**, **PRIMME** — and it adds
two things they don't provide:

1. **Numerical certificates on every result.** Each call returns residuals, a
   backward-error bound, orthogonality loss, and a `passed`/`failed` flag — and
   sets `passed = FALSE` rather than assert a bound it can only estimate.
2. **A structure-aware planner.** Dense, sparse-CSC, diagonal, scaled, centered,
   and composed operators route through native C++ kernels *without densifying*,
   and the planner records which kernel ran.

For the problem types above it is also faster than RSpectra, irlba, and PRIMME
(see [Benchmarks](#benchmarks)). Outside them, it says so and provides drop-in
wrappers so you can fall back per call.

## Installation

``` r
# install.packages("pak")
pak::pak("bbuchsbaum/eigencore")
```

## Quick start

A common starting point: the top few singular triplets of a tall sparse matrix —
the computation behind sparse PCA, LSA, and low-rank approximations.

```{r quick-start}
library(eigencore)
library(Matrix)

set.seed(2)
A <- as(rsparsematrix(100000, 500, density = 0.002), "dgCMatrix")

fit <- svd_partial(A, rank = 10, target = largest())
fit
```

The output shows which kernel ran, how many triplets converged, the worst
residual / backward error / orthogonality loss across the returned basis, and
whether the certificate passed. Here the norm bound is exact (`frobenius_exact`),
and the solve is faster than each baseline — 1.6× RSpectra, 9.5× irlba, 2.7×
PRIMME (see [Benchmarks](#benchmarks)).

```{r scree, echo = FALSE, fig.alt = "The ten largest singular values highlighted in blue against the full 500-point singular spectrum of A in grey."}
sig_all <- sort(sqrt(pmax(0, eigen(as.matrix(crossprod(A)),
                                   symmetric = TRUE, only.values = TRUE)$values)),
                decreasing = TRUE)
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(sig_all), sig_all, pch = 19, cex = 0.4, col = "grey78",
     bty = "n", xlab = "rank (largest to smallest)", ylab = "singular value")
points(seq_len(10), sort(fit$d, decreasing = TRUE),
       pch = 19, cex = 1.3, col = "#2166ac")
legend("topright", c("full spectrum (dense reference)", "computed by eigencore"),
       pch = 19, pt.cex = c(0.7, 1.3), col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Certificates

RSpectra, irlba, and PRIMME return vectors and values. eigencore also attaches a
certificate covering *both* singular relations (`||A v - sigma u||` and
`||A^T u - sigma v||`), and reports when the bound is only an estimate rather
than an exact guarantee.

On the problem above the certificate passes, with an exact norm bound:

```{r certificate}
fit$certificate
```

For a **column-centered** sparse matrix eigencore solves without forming the
dense centered copy — but the only norm bound available cheaply for that operator
is a *stochastic estimate*, so it returns singular values but sets
`passed = FALSE`, with a note explaining why:

```{r withheld}
cen <- svd_partial(center(A, columns = TRUE), rank = 5, target = largest())

cen$certificate$passed
cen$certificate$norm_bound_type
cen$certificate$notes
```

Results that other solvers return unconditionally, eigencore marks as unverified
when the bound cannot be established exactly.

## Operators without densifying

Centering a large sparse matrix normally destroys its sparsity — a dense centered
copy of `A` would occupy **400 MB**, versus a few MB for the sparse original.
eigencore's operator algebra lets you express the centred (or scaled, or
composed) map and solve it through native callbacks, so that dense matrix is
**never formed**:

```{r operators}
A_centered <- center(A, columns = TRUE)        # a 100000 x 500 operator, not a matrix
svd_partial(A_centered, rank = 5, target = largest())$d
```

Build operators with `linear_operator()`, combine them with `compose()`,
`crossprod_operator()`, `scale_cols()`, `center()`, …; the planner inspects the
resulting structure and routes accordingly. Call `plan_solver()` to see the chosen
path *before* you solve:

```{r plan}
plan_solver(svd_problem(A_centered, target = largest()), rank = 5)$method
```

## Partial eigendecomposition

The same interface applies to symmetric eigenproblems. Here is a sparse
second-difference operator (a 1-D graph Laplacian), asking for its **smallest**
eigenvalues:

```{r eigen}
n <- 20000
L <- bandSparse(n, n, k = c(-1, 0, 1),
                diagonals = list(rep(-1, n - 1), rep(2, n), rep(-1, n - 1)))
L <- as(L, "dgCMatrix")

eig <- eig_partial(L, k = 8, target = smallest())
eig
```

On this matrix eigencore is 43× faster than RSpectra. The exact spectrum is
known in closed form, so the computed eigenvalues can be compared directly:

```{r spectrum, echo = FALSE, fig.alt = "The eight smallest eigenvalues highlighted in blue at the bottom of the full analytic spectrum of the 20,000-point 1-D Laplacian shown in grey."}
lambda_all <- 4 * sin(seq_len(n) * pi / (2 * (n + 1)))^2  # exact Laplacian spectrum
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(lambda_all), lambda_all, type = "l", col = "grey78", lwd = 1.4,
     bty = "n", xlab = "rank (smallest to largest)", ylab = "eigenvalue")
points(seq_along(eig$values), sort(eig$values), pch = 19, cex = 1.2, col = "#2166ac")
legend("topleft", c("full spectrum (analytic)", "computed by eigencore"),
       lwd = c(1.4, NA), pch = c(NA, 19), pt.cex = 1.2,
       col = c("grey78", "#2166ac"), bty = "n")
par(op)
```

## Benchmarks {#benchmarks}

Median wall-clock time. eigencore's column is solve + certificate; the baseline
columns are solve only. Reproduce with `Rscript inst/benchmarks/bench-readme.R`.

| Problem (all certified `passed` by eigencore) | eigencore | vs RSpectra | vs irlba | vs PRIMME |
|---|---|---|---|---|
| Tall sparse SVD, 100000 × 500, k = 10 | 16 ms | **1.6× faster** | **9.5× faster** | **2.7× faster** |
| Wide sparse SVD, 500 × 100000, k = 10 | 14 ms | **11.4× faster** | **25.3× faster** | **12.7× faster** |
| Banded Hermitian, smallest, n = 20000, k = 8 | 67 ms | **43.4× faster** | — | — |

<sub>R 4.5.1 · aarch64-apple-darwin20 · RSpectra 0.16.2 / irlba 2.3.7 /
PRIMME 3.2.6. The tall/wide SVD rows also allocate ~6–8× less memory than irlba.
Numbers are machine- and BLAS-dependent; rerun locally for your own.</sub>

**Where eigencore is *not* the fast path.** The speedups above come from
structure-aware kernels — the certified Gram path for tall/wide sparse SVD whose
small side is ≤ 512, and the banded/tridiagonal Hermitian solver. On a general
2-D-grid sparse Laplacian, or a tall SVD whose narrow side exceeds that Gram
limit, RSpectra is currently faster. eigencore records which path ran in
`fit$method`.

## eigencore or RSpectra / irlba?

**Reach for eigencore when** you need certificates on results; you have
tall/wide sparse or PCA-shaped SVD problems (faster than RSpectra and PRIMME,
less memory than irlba); you need the smallest eigenvalues of a
banded/structured Hermitian operator; or you want to centre, scale, or compose
operators without forming a dense matrix.

**Stay with RSpectra / irlba when** you have a general large sparse problem
outside eigencore's promoted kernels and don't need certificates — they are
mature and fast. eigencore's compatible wrappers (`eigs()`, `eigs_sym()`,
`svds()`) let you switch back with a one-line change.

## RSpectra-compatible drop-in

If you already call `RSpectra::eigs_sym()`, swapping in `eigencore::eigs_sym()` is
a one-line change that also returns a certificate:

```{r rspectra-shim}
res <- eigs_sym(L, k = 8, which = "SA")
res$values
res$certificate$passed
```

`eigs()`, `eigs_sym()`, and `svds()` accept the same `which` codes as `RSpectra`
(`"LM"`, `"SM"`, `"LA"`, `"SA"`, `"LR"`, `"SR"`, `"LI"`, `"SI"`, `"BE"`).

## What makes eigencore different

- **Certificates, withheld when the bound is only an estimate.** Every result
  carries `max_residual`, `max_backward_error`, `max_orthogonality_loss`, and a
  labeled norm bound (`frobenius_exact`, `frobenius_hutchinson_estimate`, …).
  `passed` is set only when the bound is exact.
- **Transparent method selection.** `fit$method` and `plan_solver()` name the
  path that actually ran — `native certified Gram SVD special case`,
  `native tridiagonal Hermitian LAPACK selected eigensolver`, or, when a problem
  class has no production kernel yet, an explicit `reference`/`oracle` label.
  Sparse inputs do not silently densify.
- **Operator algebra over a native engine.** Dense, CSC, diagonal, scaled,
  centered, and composed operators apply to dense blocks through a C++17 ABI
  with no R-level iteration in the hot loop.
- **BLAS-level implementation throughout.** Certificate Gram products use
  `dsyrk` rather than `dgemm`, halving the flop count. Arnoldi refined-Ritz
  extraction forms Ritz vectors with two `dgemv` calls per eigenpair (3×
  faster at n = 10⁵ vs the prior scalar loop). Lanczos reorthogonalization
  uses block CGS2 (`dgemv`), matching the Golub-Kahan path.

Run `vignette("eigencore", package = "eigencore")` for a guided tour and
`vignette("certificates", package = "eigencore")` for the deep dive on reading the
numerical evidence — including what to do when a check fails.

## Status

**eigencore is pre-1.0: the API is still settling and is not yet on CRAN.** Treat
the public surface as stabilising, not frozen.

That caveat is about *lifecycle*, not quality. The V2 release surface is
tested — each promoted path has benchmark evidence against RSpectra and PRIMME,
an adversarial test suite covers the solvers and their certificates, and every
result carries a certificate. Broader general-sparse and matrix-free SVD, scalable
interior workflows, and full nonsymmetric Krylov–Schur extraction remain V3 scope,
and any path without a production kernel carries an honest `reference`/`oracle`
label. See
[`docs/method-selection-and-workflows.md`](docs/method-selection-and-workflows.md)
for the workflow map,
[`docs/v1-benchmark-manifest.md`](docs/v1-benchmark-manifest.md) for the
benchmark-gate inventory, and
[`docs/known-limitations.md`](docs/known-limitations.md) for current boundaries.

## License

MIT © Bradley Buchsbaum.
