---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 100
)
```

# eigencore

**eigencore** is a native engine for certified partial eigenvalue and singular-value
computation in R. The goal is a partial spectrum that is **fast enough to use
in production and trustworthy enough to build on** — every returned result
carries residuals, backward-error estimates, orthogonality loss, and an
inspectable solver plan that records exactly which kernel ran. The package
ships RSpectra-compatible shims (`eigs()`, `eigs_sym()`, `svds()`) so existing
code can migrate gradually.

## Installation

eigencore is in early development. You can install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("bbuchsbaum/eigencore")
```

## Quick start

```{r quick-start}
library(eigencore)

set.seed(1)
n <- 200
A <- crossprod(matrix(rnorm(n * n), n, n)) / n + diag(n)

fit <- eig_partial(A, k = 5, target = largest())
fit
```

The result prints a one-screen ledger: which solver actually ran, how many
pairs converged, the worst residual and backward error across the returned
basis, and whether the certificate passed.

```{r certificate}
fit$certificate
```

The certificate is the differentiator — eigencore tells you whether the
numbers are trustworthy, not just what they are.

A partial decomposition is a certified *slice* of the spectrum: you compute the
part you need and leave the rest untouched.

```{r spectrum, echo = FALSE, fig.alt = "The five largest eigenvalues highlighted in blue against the full 200-point spectrum of A in grey."}
all_vals <- sort(eigen(A, symmetric = TRUE, only.values = TRUE)$values,
                 decreasing = TRUE)
op <- par(mar = c(4, 4.4, 1, 1))
plot(seq_along(all_vals), all_vals, pch = 19, cex = 0.4, col = "grey78",
     bty = "n", xlab = "rank (largest to smallest)", ylab = "eigenvalue")
points(seq_along(fit$values), sort(fit$values, decreasing = TRUE),
       pch = 19, cex = 1.3, col = "#2166ac")
legend("topright", c("full spectrum (dense reference)", "computed by eigencore"),
       pch = 19, pt.cex = c(0.7, 1.3), col = c("grey78", "#2166ac"),
       bty = "n")
par(op)
```

## RSpectra-compatible drop-in

If you have existing code calling `RSpectra::eigs_sym()`, swapping in
`eigencore::eigs_sym()` is a one-line change and gives you a certificate for free:

```{r rspectra-shim}
res <- eigs_sym(A, k = 5, which = "LA")
res$values
res$certificate
```

`eigs()`, `eigs_sym()`, and `svds()` accept the same `which` codes as
`RSpectra` (`"LM"`, `"SM"`, `"LA"`, `"SA"`, `"LR"`, `"SR"`, `"LI"`, `"SI"`,
`"BE"`).

## Partial SVD

```{r svd}
M <- matrix(rnorm(400 * 50), 400, 50)
svd_fit <- svd_partial(M, rank = 5, target = largest())
svd_fit
```

The `method` field reports which path ran. eigencore takes a fast certified
Gram special case when it can prove the result to tolerance, and falls back to a
Golub–Kahan bidiagonalization when the Gram route can't be certified. Either
way the certificate covers *both* singular relations, `||A v - sigma u||` and
`||A^T u - sigma v||`, so the triplets come back with honest backward-error
bounds — eigencore never takes a shortcut it can't certify.

## What makes eigencore different

- **Certificates by default.** Every result carries `max_residual`,
  `max_backward_error`, `max_orthogonality_loss`, and a `passed`/`failed`
  verdict. The returned norm bound is labeled (`frobenius_exact`,
  `frobenius_hutchinson_estimate`, ...) so you know how the certificate was
  computed.
- **Planner honesty.** The `method` field on every result identifies the path
  that actually ran — `native scalar thick-restart Hermitian Lanczos`,
  `native certified Gram SVD special case`, `native dense LAPACK SVD fallback`,
  or, when a problem class has no production kernel yet,
  `reference Hermitian Lanczos (prototype/oracle fallback)`. Sparse inputs do
  not silently densify; reference oracles are never dressed as production paths.
  Call `plan_solver()` to see the chosen path *before* you solve.
- **Block-native engine.** Operators apply to dense blocks through a frozen
  C++17 ABI. Dense, CSC, diagonal, scaled, centered, and composed operators
  go through native kernels without R-level iteration in the hot loop.
- **First-class operator algebra.** Build operators with `linear_operator()`,
  combine them with `compose()`, `crossprod_operator()`, `scale_cols()`,
  `center()`, etc. — the planner inspects the resulting structure and routes
  accordingly.

Run `vignette("eigencore", package = "eigencore")` for a guided tour of
operators, problems, plans, and certificates, and
`vignette("certificates", package = "eigencore")` for the deep dive on reading
the numerical evidence — including what to do when a check fails.

## Status

**eigencore is experimental.** The public surface is settling but now has a
scoped V1 release surface with fresh local gate evidence. The V1 native engine
ships scalar Hermitian Lanczos plus explicit and diagnostic block Hermitian
Lanczos, native structured-tridiagonal Hermitian defaults, scoped sparse
generalized LOBPCG, certified tall/wide sparse SVD special cases, scoped
randomized SVD acceleration, dense LAPACK fallbacks, dense/structured
shift-invert, and dense/sparse-CSC nonsymmetric Arnoldi compatibility. Broader
general sparse and matrix-free SVD, fully native randomized solver control,
general sparse native LU, matrix-free native Arnoldi restart, and native/block
generalized Lanczos remain future scope. See
[`plan_v1.md`](plan_v1.md) and the docs under `docs/` for the engineering
roadmap and current limitations. Start with
[`docs/method-selection-and-workflows.md`](docs/method-selection-and-workflows.md)
for the current API workflow map and
[`docs/v1-benchmark-manifest.md`](docs/v1-benchmark-manifest.md) for the
benchmark gate inventory.

## License

MIT © Bradley Buchsbaum.
