---
title: Benchmarking eigencore against established solvers
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Benchmarking eigencore against established solvers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.6,
  fig.align = "center",
  out.width = "96%",
  dpi = 100
)

benchmark_ready <- all(vapply(
  c("bench", "RSpectra", "irlba"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))
benchmark_iterations <- 3L
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r helpers, include = FALSE}
`%||%` <- function(x, y) if (is.null(x)) y else x

fro_norm <- function(A) {
  if (inherits(A, "sparseMatrix")) {
    sqrt(sum(A@x^2))
  } else {
    sqrt(sum(abs(A)^2))
  }
}

relative_error <- function(x, truth) {
  x <- sort(Re(x), decreasing = TRUE)
  truth <- sort(Re(truth), decreasing = TRUE)
  max(abs(x - truth) / pmax(1, abs(truth)))
}

eigen_backward_error <- function(A, values, vectors) {
  A_dense <- as.matrix(A)
  values <- Re(values)
  vectors <- as.matrix(vectors)
  residual <- A_dense %*% vectors - vectors %*% diag(values, nrow = length(values))
  scale <- fro_norm(A_dense) + abs(values)
  max(sqrt(colSums(abs(residual)^2)) / pmax(scale, .Machine$double.eps))
}

svd_backward_error <- function(A, d, u, v) {
  A_dense <- as.matrix(A)
  d <- Re(d)
  u <- as.matrix(u)
  v <- as.matrix(v)
  left_residual <- A_dense %*% v - u %*% diag(d, nrow = length(d))
  right_residual <- t(A_dense) %*% u - v %*% diag(d, nrow = length(d))
  residual <- sqrt(colSums(abs(left_residual)^2) + colSums(abs(right_residual)^2))
  max(residual / pmax(fro_norm(A_dense) + d, .Machine$double.eps))
}

bench_eval <- function(expr, iterations = benchmark_iterations) {
  expr <- substitute(expr)
  env <- parent.frame()
  last_result <- NULL
  mark <- tryCatch(
    bench::mark(
      last_result <- eval(expr, env),
      iterations = iterations,
      check = FALSE,
      memory = TRUE,
      filter_gc = FALSE
    ),
    error = function(e) e
  )
  if (inherits(mark, "error")) {
    return(list(
      result = NULL,
      median_ms = NA_real_,
      mem_mb = NA_real_,
      error = conditionMessage(mark)
    ))
  }
  list(
    result = last_result,
    median_ms = as.numeric(mark$median[[1L]]) * 1000,
    mem_mb = as.numeric(mark$mem_alloc[[1L]]) / 1024^2,
    error = NA_character_
  )
}

make_dense_hermitian <- function(n, seed = 1L) {
  set.seed(seed)
  X <- matrix(rnorm(n * n), n, n)
  crossprod(X) / n + diag(seq(1, 1.2, length.out = n))
}

make_dense_low_rank <- function(m, n, rank = 8L, noise = 1e-3, seed = 1L) {
  set.seed(seed)
  U <- qr.Q(qr(matrix(rnorm(m * rank), m, rank)))
  V <- qr.Q(qr(matrix(rnorm(n * rank), n, rank)))
  signal <- U %*% diag(seq(rank, 1, length.out = rank), nrow = rank) %*% t(V)
  signal + noise * matrix(rnorm(m * n), m, n)
}

path_laplacian <- function(n) {
  Matrix::bandSparse(
    n,
    k = c(-1L, 0L, 1L),
    diagonals = list(rep(-1, n - 1L), c(1, rep(2, n - 2L), 1), rep(-1, n - 1L))
  )
}

eigen_rows <- function(name, A, k = 6L, tol = 1e-8, seed = 1L,
                       iterations = benchmark_iterations) {
  truth <- eigen(as.matrix(A), symmetric = TRUE, only.values = TRUE)$values[seq_len(k)]
  methods <- c("eigencore", "RSpectra", "base")
  rows <- lapply(methods, function(method) {
    timed <- switch(
      method,
      eigencore = bench_eval({
        set.seed(seed)
        eig_partial(A, k = k, target = largest(), tol = tol)
      }, iterations = iterations),
      RSpectra = bench_eval({
        RSpectra::eigs_sym(A, k = k, which = "LA", opts = list(tol = tol, maxitr = 1000L))
      }, iterations = iterations),
      base = bench_eval({
        eigen(as.matrix(A), symmetric = TRUE)
      }, iterations = iterations)
    )
    if (!is.na(timed$error)) {
      return(data.frame(
        regime = name,
        task = "eigen",
        method = method,
        median_ms = timed$median_ms,
        mem_mb = timed$mem_mb,
        rel_error = NA_real_,
        backward_error = NA_real_,
        residual_check = FALSE,
        eigencore_label = NA_character_,
        status = timed$error,
        stringsAsFactors = FALSE
      ))
    }
    result <- timed$result
    extracted <- switch(
      method,
      eigencore = list(values = values(result), vectors = vectors(result)),
      RSpectra = list(values = result$values, vectors = result$vectors),
      base = {
        ord <- order(result$values, decreasing = TRUE)[seq_len(k)]
        list(values = result$values[ord], vectors = result$vectors[, ord, drop = FALSE])
      }
    )
    backward <- eigen_backward_error(A, extracted$values, extracted$vectors)
    data.frame(
      regime = name,
      task = "eigen",
      method = method,
      median_ms = timed$median_ms,
      mem_mb = timed$mem_mb,
      rel_error = relative_error(extracted$values, truth),
      backward_error = backward,
      residual_check = isTRUE(backward <= tol),
      eigencore_label = if (method == "eigencore") result$method else NA_character_,
      status = "ok",
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

svd_rows <- function(name, A, rank = 6L, tol = 1e-8, seed = 1L,
                     iterations = benchmark_iterations) {
  truth <- svd(as.matrix(A), nu = 0, nv = 0)$d[seq_len(rank)]
  methods <- c("eigencore", "RSpectra", "irlba", "base")
  rows <- lapply(methods, function(method) {
    timed <- switch(
      method,
      eigencore = bench_eval({
        set.seed(seed)
        svd_partial(A, rank = rank, target = largest(), tol = tol)
      }, iterations = iterations),
      RSpectra = bench_eval({
        RSpectra::svds(A, k = rank, nu = rank, nv = rank, opts = list(tol = tol, maxitr = 1000L))
      }, iterations = iterations),
      irlba = bench_eval({
        set.seed(seed)
        irlba::irlba(A, nv = rank, nu = rank, tol = tol)
      }, iterations = iterations),
      base = bench_eval({
        svd(as.matrix(A), nu = rank, nv = rank)
      }, iterations = iterations)
    )
    if (!is.na(timed$error)) {
      return(data.frame(
        regime = name,
        task = "SVD",
        method = method,
        median_ms = timed$median_ms,
        mem_mb = timed$mem_mb,
        rel_error = NA_real_,
        backward_error = NA_real_,
        residual_check = FALSE,
        eigencore_label = NA_character_,
        status = timed$error,
        stringsAsFactors = FALSE
      ))
    }
    result <- timed$result
    extracted <- switch(
      method,
      eigencore = list(d = values(result), u = left_vectors(result), v = right_vectors(result)),
      RSpectra = list(d = result$d, u = result$u, v = result$v),
      irlba = list(d = result$d, u = result$u, v = result$v),
      base = list(d = result$d[seq_len(rank)], u = result$u, v = result$v)
    )
    backward <- svd_backward_error(A, extracted$d, extracted$u, extracted$v)
    data.frame(
      regime = name,
      task = "SVD",
      method = method,
      median_ms = timed$median_ms,
      mem_mb = timed$mem_mb,
      rel_error = relative_error(extracted$d, truth),
      backward_error = backward,
      residual_check = isTRUE(backward <= tol),
      eigencore_label = if (method == "eigencore") result$method else NA_character_,
      status = "ok",
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

plot_time_summary <- function(rows) {
  ok <- rows[rows$status == "ok", , drop = FALSE]
  labels <- paste(ok$regime, ok$method, sep = "\n")
  colors <- c(eigencore = "#2166ac", RSpectra = "#b2182b", irlba = "#1b7837", base = "grey55")
  op <- par(mar = c(9, 4.5, 1, 1))
  on.exit(par(op))
  barplot(
    ok$median_ms,
    names.arg = labels,
    las = 2,
    col = colors[ok$method],
    border = NA,
    ylab = "median elapsed time (ms)",
    cex.names = 0.68
  )
  legend("topright", names(colors), fill = colors, border = NA, bty = "n", cex = 0.85)
}
```

Benchmarks are trust work. They tell you whether a solver is fast enough,
whether it returns the same numerical slice as established tools, and whether
the returned vectors actually satisfy the equations they claim to solve.

This vignette runs a compact comparison suite against `RSpectra`, base R, and
`irlba`. It uses small problems so the article can be rebuilt on a laptop.
Treat the tables as a reproducible smoke comparison, not as a release-speed
claim. For promotion-grade evidence, use the installed-package benchmark gates
at the end of the vignette.

```{r setup}
library(eigencore)
```

## What should be compared?

A useful benchmark suite needs more than one matrix. Dense and sparse inputs
exercise different planner decisions, and SVD has different baselines than
Hermitian eigenproblems. The smoke suite below covers five regimes:

```{r regimes}
benchmark_regimes <- data.frame(
  regime = c(
    "dense Hermitian",
    "sparse path Laplacian",
    "dense low-rank SVD",
    "tall sparse SVD",
    "wide sparse SVD"
  ),
  input = c("120 x 120 dense", "300 x 300 dgCMatrix", "180 x 70 dense",
            "320 x 60 dgCMatrix", "60 x 320 dgCMatrix"),
  compared_methods = c("eigencore, RSpectra, base",
                       "eigencore, RSpectra, base",
                       "eigencore, RSpectra, irlba, base",
                       "eigencore, RSpectra, irlba, base",
                       "eigencore, RSpectra, irlba, base")
)
knitr::kable(benchmark_regimes)
```

Base R is included as a truth-oriented reference. For sparse matrices it
intentionally densifies via `as.matrix()`, so it is useful at small sizes but
not a fair large-sparse production baseline. `irlba` is included for SVD rows,
where it is a standard partial-SVD comparator.

## How are the rows judged?

Each row records four things:

- median elapsed time from `bench::mark()`;
- allocated memory reported by `bench`;
- relative error against a dense base truth for the requested values;
- an explicit residual/backward-error check on the returned vectors.

For eigencore rows, the table also reports the planner label that ran. That
label matters: a native row, a dense fallback, and a reference boundary are
different claims even if their values agree.

## Run the smoke comparison

```{r build-cases, eval = benchmark_ready, include = FALSE}
set.seed(1001)
bench_cases <- list(
  dense_hermitian = make_dense_hermitian(120L, seed = 1001L),
  sparse_laplacian = path_laplacian(300L),
  dense_low_rank_svd = make_dense_low_rank(180L, 70L, rank = 8L, seed = 1002L),
  tall_sparse_svd = Matrix::rsparsematrix(320L, 60L, density = 0.035),
  wide_sparse_svd = Matrix::rsparsematrix(60L, 320L, density = 0.035)
)
```

```{r run-benchmarks, eval = benchmark_ready, include = FALSE}
benchmark_rows <- rbind(
  eigen_rows("dense Hermitian", bench_cases$dense_hermitian,
             k = 6L, seed = 2001L),
  eigen_rows("sparse path Laplacian", bench_cases$sparse_laplacian,
             k = 6L, seed = 2002L),
  svd_rows("dense low-rank SVD", bench_cases$dense_low_rank_svd,
           rank = 6L, seed = 2003L),
  svd_rows("tall sparse SVD", bench_cases$tall_sparse_svd,
           rank = 6L, seed = 2004L),
  svd_rows("wide sparse SVD", bench_cases$wide_sparse_svd,
           rank = 6L, seed = 2005L)
)
stopifnot(all(is.finite(benchmark_rows$median_ms[benchmark_rows$status == "ok"])))
stopifnot(all(is.finite(benchmark_rows$rel_error[benchmark_rows$status == "ok"])))
stopifnot(all(is.finite(benchmark_rows$backward_error[benchmark_rows$status == "ok"])))
```

```{r benchmark-table, eval = benchmark_ready, echo = FALSE}
display_rows <- benchmark_rows
display_rows$median_ms <- signif(display_rows$median_ms, 4)
display_rows$mem_mb <- signif(display_rows$mem_mb, 4)
display_rows$rel_error <- signif(display_rows$rel_error, 3)
display_rows$backward_error <- signif(display_rows$backward_error, 3)
knitr::kable(
  display_rows[, c("regime", "task", "method", "median_ms", "mem_mb",
                   "rel_error", "backward_error", "residual_check",
                   "eigencore_label", "status")],
  caption = paste("Smoke benchmark rows with", benchmark_iterations,
                  "iterations per method.")
)
```

```{r benchmark-plot, eval = benchmark_ready, echo = FALSE, fig.cap = "Elapsed-time smoke comparison by regime and method. These small rows are useful for detecting obvious regressions and method-selection surprises, but they are not release-grade speed evidence.", fig.alt = "Bar chart of median elapsed time in milliseconds by benchmark regime and solver method."}
plot_time_summary(benchmark_rows)
```

Read the table from right to left. First, `status` should be `"ok"` and
`residual_check` should be `TRUE`; otherwise the elapsed time is not a useful
time-to-answer measurement. Next, `rel_error` should be near machine precision
or solver tolerance. Only then does `median_ms` become meaningful.

For eigencore, inspect `eigencore_label` before interpreting speed. A native
sparse path and a dense LAPACK fallback both may be correct, but they support
different scaling claims.

## What this smoke vignette does not prove

The rows above are deliberately small. They keep the vignette runnable, and
they catch obvious problems: wrong target, broken sparse handling, missing
vectors, certification drift, and large timing regressions. They do not prove
large-scale speed claims.

For larger sparse matrices, base R is a truth oracle only while `as.matrix(A)`
is affordable. The production question is whether eigencore beats certified
partial solvers without silently densifying. That requires installed-package
runs, larger fixtures, saved artifacts, and memory evidence.

## Run promotion-grade benchmarks

Use the `inst/benchmarks` scripts when you need evidence suitable for a release
note, paper, or solver-promotion decision. Run them from an installed package,
not from a mutable source tree:

```sh
R CMD INSTALL --library=/tmp/eigencore-bench-lib .
R_LIBS=/tmp/eigencore-bench-lib \
  Rscript inst/benchmarks/bench-native-hermitian-gate.R \
  --strict --save --cases=path_laplacian:1000
```

For the SVD surface, compare eigencore against `RSpectra`, `irlba`, and base
rows where those baselines are meaningful:

```sh
R_LIBS=/tmp/eigencore-bench-lib \
  Rscript inst/benchmarks/bench-svd-surface.R \
  --iterations=1 --h-candidate \
  --methods=eigencore,RSpectra,irlba,base \
  --cases=tall_sparse,wide_sparse \
  --subject=eigencore --strict --save
```

Those scripts add the release-grade pieces this vignette intentionally keeps
small: strict pass/fail gates, larger sparse fixtures, saved RDS artifacts,
memory ratios, and planner-provenance checks. Use
`docs/v1-benchmark-manifest.md` for the current release-gate inventory and
`docs/post-v1-benchmark-gates.md` for future-promotion gates.
