---
title: 'Certificates: reading the numerical evidence'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Certificates: reading the numerical evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.1,
  fig.align = "center",
  out.width = "92%",
  dpi = 100
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r helpers, include = FALSE}
# Shared plotting helper. Kept out of the reader's view: the reader sees
# the API call and the picture, never the plotting boilerplate.
ec_blue <- "#2166ac"; ec_red <- "#b2182b"; ec_tol <- "#ef8a62"

# Per-pair backward error as a stem/dot plot, against the tolerance line.
# Converged pairs are blue, failed pairs red.
plot_backward_error <- function(cert, main = NULL, tol = cert$tolerance,
                                ylim = NULL) {
  be <- pmax(cert$backward_error, .Machine$double.eps)
  k  <- length(be)
  col_pt <- ifelse(cert$converged, ec_blue, ec_red)
  if (is.null(ylim)) {
    rng  <- range(c(be, tol))
    ylim <- c(rng[1] / 6, rng[2] * 6)
  }
  op <- par(mar = c(4, 4.6, if (is.null(main)) 1 else 2.4, 1)); on.exit(par(op))
  plot(seq_len(k), be, log = "y", type = "n", xlim = c(0.5, k + 0.5),
       ylim = ylim, xaxt = "n", bty = "n", main = main,
       xlab = "returned pair", ylab = "backward error (log scale)")
  axis(1, at = seq_len(k))
  segments(seq_len(k), ylim[1], seq_len(k), be, col = col_pt, lwd = 2)
  points(seq_len(k), be, pch = 19, cex = 1.3, col = col_pt)
  abline(h = tol, col = ec_tol, lty = 2, lwd = 2)
  text(k + 0.45, tol, sprintf("tol = %.0e", tol), col = ec_tol,
       pos = 3, cex = 0.85, xpd = NA)
}
```

Every result returned by eigencore carries a **certificate**: a compact record
of the residual, backward-error scale, orthogonality check, and tolerance used
for that result. The certificate lets you inspect those checks without
recomputing the residual. It does not certify spectral ordering or any property
that its `certificate_type` does not name.

This vignette walks through the certificate's fields, the three things they
measure, and what to do when one of them fails.

```{r setup}
library(eigencore)
```

## Anatomy of a certificate

```{r anatomy}
set.seed(1)
n <- 200
A <- crossprod(matrix(rnorm(n * n), n, n)) / n + diag(n)

fit  <- eig_partial(A, k = 5, target = largest())
cert <- fit$certificate
cert
```

The fields you act on, in order:

- **`passed`** — overall verdict. `TRUE` means every returned pair satisfies
  the checks required by this certificate type. The backward-error scale is
  exact or a lower bound on the matrix 2-norm, never an estimate, so a pass
  is never optimistic (see "Norm bounds" below).
- **`tolerance`** — the user-requested tolerance. Defaults to `1e-8`.
- **`max_residual`** — the worst absolute residual `||A v - lambda v||` (or
  `||A v - lambda B v||` for generalized problems) across the returned basis.
- **`max_backward_error`** — the worst residual divided by the labelled scale.
  This is the number that has to be below `tolerance` for a pair to be
  considered converged.
- **`max_orthogonality_loss`** — the worst entry of `|V* V - I|` (or
  `|V* B V - I|` in B-inner-product problems).
- **`failed_indices`** — which pairs failed. Empty when `passed = TRUE`.

Three more fields explain *how* the verdict was reached rather than what it
is: `norm_bound_type` says whether the `||A||_2` (and `||B||_2`) value in the
backward-error scale is exact or a lower bound, `norm_source` says where it
came from, and `norm_values` holds the values used. `scale_is_estimate` is
always `FALSE` for built-in certificates. `certificate_type` and `notes`
carry provenance text for unusual states. `?certificate` documents every
field.

The `max_*` fields are summaries. The certificate also keeps the underlying
*per-pair* vectors (`cert$backward_error`, `cert$converged`), so you can see at
a glance whether the whole basis cleared the bar or just barely missed:

```{r anatomy-bars, echo = FALSE, fig.cap = "Per-pair backward error for the five returned eigenpairs. The `max_backward_error` field is simply the height of the tallest stem; `passed` is TRUE because every stem clears the tolerance line.", fig.alt = "Stem plot of backward error for five eigenpairs on a log scale, all falling well below the dashed tolerance line."}
plot_backward_error(cert)
```

## Three things a certificate measures

### 1. Residual

For a Hermitian eigenproblem `A v = lambda v`, the *absolute* residual is
`||A v_i - lambda_i v_i||`. For an SVD, both the left and right relations
matter, so the combined residual is

> sqrt( ||A v - sigma u||^2  +  ||A^T u - sigma v||^2 ).

For a generalized SPD problem `A v = lambda B v`, the residual is computed
in the *original* coordinates: `||A v_i - lambda_i B v_i||`. Eigencore never
reports a residual computed in a transformed problem space without
saying so in `certificate_type`.

### 2. Backward error

Absolute residuals can be misleading when the operator's norm is large or
small. Backward error is the residual divided by a scale that captures
"how big could a perturbation of `A` (and `B`) be that *exactly* makes
`(lambda, v)` an eigenpair?". eigencore uses the standard normwise definition
in the matrix 2-norm:

> eta_i  =  ||r_i||  /  ( (||A||_2 + |lambda_i| ||B||_2) ||v_i|| ),

and for a singular triplet `eta_i = sqrt(||A v - sigma u||^2 + ||A^T u - sigma
v||^2) / ||A||_2` (unit vectors). A pair "converges" when `eta_i <= tol`. The
tolerance you pass to `eig_partial(tol = ...)` is this backward-error
tolerance, not a residual tolerance.

`||A||_2` is rarely known exactly, so the denominator uses a value that never
exceeds it. A smaller denominator can only make the reported `eta_i` larger
than the true backward error, so a certificate that passes would also pass
with the exact norm. (Before eigencore 1.4 the scale was the Frobenius norm,
which can be up to `sqrt(rank)` times larger than `||A||_2` and made the check
correspondingly more lenient.)

### 3. Orthogonality

Iterative methods drift. After enough restarts the returned basis can lose
orthogonality even when each per-pair residual is small. The certificate
records `max_abs(V* V - I)` (or `max_abs(V* B V - I)`); its acceptance
tolerance is `max(tol, sqrt(eps))`. A passing residual should be
interpreted alongside the orthogonality field. If orthogonality loss exceeds
its tolerance, clustered or repeated eigenvalues may be represented by nearly
duplicate vectors even when individual residuals are small.

A certificate evaluates the pairs the solver returned. It does not prove that
those pairs are the requested largest, smallest, or nearest eigenvalues. When
target identity is consequential, inspect the method and target in the plan
and compare against an independent oracle on a tractable instance.

## Reading common certificate states

The single most useful habit is to picture the per-pair backward error against
the tolerance line. A passing certificate is "all stems below the line"; a
failing one is "at least one stem above it." Here are the same ten eigenpairs
of `A`, computed two ways:

```{r pass-vs-fail-fits}
# Largest eigenvalues are well separated -> easy, converges fast.
fit_pass <- eig_partial(A, k = 10, target = largest())

# Smallest eigenvalues are densely clustered near 1 -> a tight maxit (restart)
# budget leaves them short of tolerance.
fit_fail <- eig_partial(A, k = 10, target = smallest(), maxit = 4)

c(largest_passed  = fit_pass$certificate$passed,
  smallest_passed = fit_fail$certificate$passed)
```

```{r pass-vs-fail, echo = FALSE, fig.width = 9, fig.height = 4, fig.cap = "Same matrix, same k, two targets. Left: the ten largest eigenpairs all clear the tolerance (blue, passed). Right: the ten smallest stall above it under a tight iteration budget (red, failed).", fig.alt = "Two side-by-side stem plots of per-pair backward error. Left panel shows ten blue points below the tolerance line; right panel shows ten red points above the tolerance line."}
op <- par(mfrow = c(1, 2))
ylim <- c(1e-16, 1e-1)
plot_backward_error(fit_pass$certificate, main = "largest(): passed", ylim = ylim)
plot_backward_error(fit_fail$certificate, main = "smallest(), maxit = 4: failed", ylim = ylim)
par(op)
```

### Clean: every box ticked

```{r clean}
fit_pass$certificate$passed
fit_pass$certificate$norm_bound_type
fit_pass$certificate$scale_is_estimate
```

This is the easy case. Every residual is below tolerance and orthogonality is
near machine precision. The scale is a lower bound on `||A||_2`, here the
largest computed Ritz value (`norm_source` is `"ritz+identity"`), which for a
largest-eigenvalue target is essentially the exact norm.

### Failed: residual too large

The right-hand panel above is a genuine failure. The smallest eigenvalues of
`A` sit in a dense cluster just above 1, so they need many more iterations than
the well-separated largest ones. With `maxit = 4` the solver runs out of
restart budget before all pairs converge, and the verdict flips. The `failed_indices`
slot tells you which Ritz pairs missed:

```{r failed}
fit_fail$certificate$passed
fit_fail$certificate$failed_indices
fit_fail$certificate$max_backward_error
```

You do not have to guess how far off it was, or whether it was inching toward
convergence. The solver records a `convergence_history`; plotting the worst
backward error per restart shows the failed run plateauing above the line while
a generous budget drives it underneath.

```{r convergence-fits}
fit_ok <- eig_partial(A, k = 10, target = smallest(), maxit = 40)
fit_ok$certificate$passed
```

```{r convergence, echo = FALSE, fig.cap = "Worst backward error per restart for the ten smallest eigenpairs. A tight budget (red) stalls above the tolerance; a generous one (blue) drives the error under the line and the certificate passes.", fig.alt = "Line plot on a log scale of backward error versus restart number. The red maxit-4 curve plateaus above the tolerance line; the blue maxit-40 curve descends below it."}
h_fail <- diagnostics(fit_fail)$convergence_history
h_ok   <- diagnostics(fit_ok)$convergence_history
tol <- fit_fail$certificate$tolerance
yl  <- range(c(h_fail$max_backward_error, h_ok$max_backward_error, tol))
op <- par(mar = c(4, 4.6, 1, 1))
tryCatch({
  plot(h_fail$restart, h_fail$max_backward_error, log = "y", type = "l",
       lwd = 2, col = ec_red, bty = "n", ylim = yl,
       xlab = "restart", ylab = "max backward error (log scale)")
  lines(h_ok$restart, h_ok$max_backward_error, lwd = 2, col = ec_blue)
  abline(h = tol, col = ec_tol, lty = 2, lwd = 2)
  legend("right", c("maxit = 4 (failed)", "maxit = 40 (passed)", "tolerance"),
         col = c(ec_red, ec_blue, ec_tol), lwd = 2, lty = c(1, 1, 2),
         bty = "n", cex = 0.9, inset = 0.02)
}, finally = par(op))
```

What to do: increase `maxit`, raise `tol`, or — when you suspect a clustered
spectrum — request a larger `k` (so the cluster is fully covered by the
returned basis) and slice afterwards. Here, lifting `maxit` (the restart limit) from 4 to 40 is
enough.

### Norm bounds: exact or a lower bound

`norm_bound_type` says what kind of `||A||_2` value scaled the backward error,
as `A` or `A+B` parts (`B = I` appears as `identity_exact`):

- `two_norm_exact` — the exact spectral norm: diagonal matrices
  (`max |d_i|`), a full computed Hermitian spectrum (`eig_full()` and dense
  solves that compute every eigenvalue), or a user-asserted
  `metadata$two_norm`.
- `two_norm_lower_bound` — a value that never exceeds `||A||_2`.

`norm_source` names where the value came from:

- `diagonal`, `full_spectrum`, `metadata`, `identity` — exact values.
- `column_norms` — the largest column norm `||A e_j||` (dense, sparse,
  tridiagonal storage; no operator applies).
- `applied_vectors` — the largest `||A x|| / ||x||` over the certified vectors,
  computed with the same operator applies as the residuals.
- `ritz` — the residual-corrected value `|lambda| ||B x|| / ||x|| - ||r|| /
  ||x||` when only residual norms are available.
- `frobenius_rank_bound` — `||A||_F / sqrt(min(m, n))` when an operator only
  carries Frobenius metadata.
- `lanczos` — a 24-step Lanczos estimate on `A` (Hermitian) or `A^T A`,
  verified as `||A y|| / ||y||` for its Ritz vector. It runs only when some
  pair fails with the cheaper bounds but could pass with a larger one, uses a
  fixed start vector (it never touches the R random-number stream), and is
  memoised per operator.

Because every source is a lower bound, matrix-free operators certify like any
other operator. A `linear_operator()` with no norm metadata used to fall back
to a stochastic Hutchinson estimate and withhold `passed`; it now certifies
with the bound from its own vectors:

```{r withheld}
set.seed(2)
op <- linear_operator(
  dim = c(n, n),
  apply = function(X, alpha = 1, beta = 0, Y = NULL) {
    Z <- alpha * (A %*% X)
    if (is.null(Y) || beta == 0) Z else Z + beta * Y
  },
  apply_adjoint = function(X, alpha = 1, beta = 0, Y = NULL) {
    Z <- alpha * (A %*% X)
    if (is.null(Y) || beta == 0) Z else Z + beta * Y
  },
  structure = hermitian(),
  name = "matrix-free Hermitian wrapper"
)
# The matrix-free path runs the reference Lanczos solver, which needs a
# larger subspace than the default to resolve this clustered spectrum.
fit_mf <- eig_partial(op, k = 5, target = largest(),
                      method = lanczos(max_subspace = 100))
fit_mf$certificate$norm_bound_type
fit_mf$certificate$norm_source
fit_mf$certificate$scale_is_estimate
fit_mf$certificate$passed
```

A lower bound can be weak — for example `||A x|| / ||x||` for the *smallest*
eigenvalues of an operator with no other information. Then the reported
backward error over-states the true one and a pair may fail although it is
accurate; the Lanczos refinement exists to close that gap before a pair is
reported as failed. What to do if a pass matters and the bound is still weak:
supply `metadata = list(two_norm = ...)` when you know `||A||_2`, or wrap a
built-in dense / `dgCMatrix` / `ddiMatrix` matrix.

### Generalized SPD: B-orthogonality matters

For `A v = lambda B v`, the certificate's orthogonality field is in the
**B-inner product**: `max_abs(V* B V - I)`. For this certificate type,
`passed = TRUE` requires both the residual threshold and B-orthogonality of the
returned vectors.

```{r generalized}
set.seed(4)
B <- diag(seq(1, 5, length.out = n))
fit_gen <- eig_partial(A, k = 5, target = largest(), B = B,
                       method = lobpcg(maxit = 200))
fit_gen$certificate$norm_bound_type
fit_gen$certificate$max_orthogonality_loss
fit_gen$certificate$passed
```

If a B-orthogonality value comes back near machine precision, the
B-inner-product Cholesky-QR refinement inside the solver did its job. If it
comes back loose (say, `1e-4`), increase `maxit` or lower `tol` — orthogonality
loss is usually the first thing to surface in ill-conditioned-B problems.

### Target completeness: the right set, not just right pairs

Residuals prove that each returned pair is an eigenpair. They cannot prove
that the returned *set* is the one you asked for. A single-vector Krylov
method sees, in exact arithmetic, only one direction of each eigenspace, so
it can miss a copy of a repeated eigenvalue: for the spectrum
`9, 9, 7, 7, 7, 5, ...` it may return `9, 9, 7, 7, 5`, and every one of
those five pairs certifies.

After a certified Hermitian Krylov solve with a `largest()`, `smallest()` or
`largest_magnitude()` target (and, with `completeness = "inertia"`, `nearest()`),
eigencore therefore checks the returned set. With an explicit dense or sparse
matrix it *counts* eigenvalues (the next section); for a matrix-free operator,
or when a factorisation would cost more than the solve, it runs a short
**deflated complement probe**: a few block Lanczos steps on the operator
restricted to the orthogonal complement of the returned vectors
(B-orthogonal for generalized problems), from a fixed-seed start that does
not touch R's random-number stream. Any Ritz value of that restricted
operator that lies beyond the least-preferred returned value (by more than
the residual and tolerance margin) proves a more-preferred eigenvalue is
missing. The solver then repairs the result with a deflated complement
solve, merges the two sets by a Rayleigh-Ritz step, re-certifies and checks
again.

```{r completeness}
d <- c(9, 9, 7, 7, 7, seq(5, 0.1, length.out = 55))
set.seed(1001)
p <- sample(60)
S <- Matrix::sparseMatrix(i = p, j = p, x = d)
bare <- eig_partial(S, k = 5, method = lanczos(completeness = "none"), seed = 1)
sort(values(bare), decreasing = TRUE)
bare$certificate$passed
fit_c <- eig_partial(S, k = 5, method = lanczos(completeness = "probe"), seed = 1)
sort(values(fit_c), decreasing = TRUE)
fit_c$certificate$target_completeness
fit_c$certificate$completeness[c("steps", "operator_columns", "rounds")]
```

The probe is cheap (by default at most 8 steps of block size 2, set by the
options `eigencore.completeness_probe_steps` and
`eigencore.completeness_probe_block`). It is a probabilistic check: it can
prove a set incomplete, not complete; an intruder whose Ritz value a short
run does not resolve goes unnoticed.

### Counting certificates

For a Hermitian matrix (or a pencil with symmetric positive definite `B`)
that is available explicitly, the number of eigenvalues below any shift `t`
is the number of negative pivots of an \(LDL^T\) factorisation of
`A - t B` (Sylvester's law of inertia). `eigen_count()` exposes that count:
dense matrices use LAPACK's Bunch-Kaufman `dsytrf`, sparse ones CHOLMOD's
\(LDL^T\) with a fill-reducing ordering, tridiagonal ones a Sturm
sequence.

```{r eigen-count}
set.seed(2)
Q <- qr.Q(qr(matrix(rnorm(60 * 60), 60)))
A7 <- Q %*% (d * t(Q))
A7 <- (A7 + t(A7)) / 2   # dense, same spectrum as S
eigen_count(S, 6)        # S is diagonal: an exact comparison
eigen_count(A7, 6)
eigen_count(A7, 7)
```

The last call lands exactly on the triple eigenvalue 7: the factorisation
reports a (numerically) zero pivot, so the shift is moved to `7 -/+ delta`,
`zero` counts the eigenvalues inside that band, and `reliable` says whether
every count came from a trustworthy factorisation. CHOLMOD's sparse
\(LDL^T\) does not pivot, so each count carries its backward-error bound
(`diagnostics$backward_bound`, from `|| |L| |D| |L'| ||`) and pivot growth;
an unreliable count is perturbed, falls back to a pivoted dense
factorisation for moderate `n`, or is reported with `reliable = FALSE`,
never silently trusted.

The completeness certificate turns two such counts into a proof. By Kahan's
theorem every returned value \(\theta_i\) has its own (distinct)
eigenvalue within \(\rho = \|R\|_F / \sigma_{\min}(X)\) of it, where
`R` holds the certificate's residuals. For a `largest()` target with
returned edge \(\theta_{\min}\), if exactly `k` eigenvalues exceed
\(\theta_{\min} - \rho - m\) (the margin `m` exceeds the
factorisation's backward error), those `k` eigenvalues are the returned
ones and the set is complete. If at least `k` eigenvalues exceed
\(\theta_{\min} + \rho + m\), a value is provably missing.

```{r inertia-certificate}
fit_i <- eig_partial(S, k = 5, method = lanczos(), seed = 1)
sort(values(fit_i), decreasing = TRUE)
fit_i$certificate$target_completeness
fit_i$certificate$completeness[c("inertia_method", "rho", "count_upper",
                                 "factorizations", "repaired")]
```

Here the first Krylov answer missed a copy of 7; the count proved it, the
repair recovered it, and the repaired set was then counted complete.

`certificate$target_completeness` is one of:

- `"inertia_verified"`: the count proves the returned set is the requested
  one (`completeness$repaired` records whether a repair was needed);
- `"inertia_failed"`: the count proves a more-preferred eigenvalue is
  missing and the repair budget
  (`options(eigencore.completeness_max_rounds = 3)`) did not fix it;
  `passed` is then `FALSE` although every residual passes
  (`residual_passed` keeps that verdict, `target_passed` is `FALSE`);
- `"inertia_inconclusive"`: the `k`-th and `(k+1)`-th eigenvalues are
  closer than the residual uncertainty (for example `k` cuts through a
  multiple eigenvalue), so the target edge is not separated; no claim is
  made either way and `passed` is unaffected;
- `"probed"`: the probe found nothing beyond the returned edge;
- `"repaired"`: the probe found a missing value and the repaired set then
  probed clean;
- `"failed"`: a probe intruder remained after the repair budget; `passed`
  is `FALSE`;
- `"exact"`: a full-spectrum route (dense LAPACK, tridiagonal, analytic grid
  Laplacian) selected from every eigenvalue, so the set is complete by
  construction;
- `"inconclusive"` (nonsymmetric probe): the probe could not reach a
  verdict (it did not converge, or the returned vectors are linearly
  dependent, as for a defective or repeated eigenvalue a single-vector
  Krylov method cannot resolve); `passed` is `FALSE`;
- `"not_checked"`: no check applied (both-ends targets, complex
  matrix-free nonsymmetric operators, a residual certificate that already
  failed, `certify = FALSE`, or `completeness = "none"`).

#### Nonsymmetric problems

Dense LAPACK routes (`dgeev`, `zgeev`, QZ for pencils) select from the full
spectrum and report `"exact"`. Krylov-Schur (Arnoldi) results, including
`eigs()`, shift-invert (probed on \((A - \sigma I)^{-1}\)), matrix-free
operators and diagonal-`B` sparse pencils (probed on \(B^{-1}A\)), are
probed: with \(Q\) an orthonormal basis of the returned vectors,
\(T = Q^T A Q\) and \(R = AQ - QT\), the eigenvalues missing from the
returned set are those of the compression
\((I - QQ^T) A (I - QQ^T)\) when \(R = 0\). A deflated Krylov-Schur run
on that compression (fixed-seed start) looks for an eigenvalue more preferred
than the returned edge by more than the margin
\(\max(\mathrm{tol}\,|\lambda|, 2\|R\|_F)\) plus its own Ritz residual;
one found is merged in by Rayleigh-Ritz on the enlarged subspace
(`"repaired"`), otherwise the result is `"failed"`. A clean probe
(`"probed"`) is evidence, not proof, and the margin covers the normal (or
well-conditioned) case only: for a non-normal matrix an eigenvalue can move
by up to its condition number times \(\|R\|\) (a Jordan block of size
`m` by \(\|R\|^{1/m}\)), which the probe does not bound. Small
complements (64 or fewer dimensions) are spanned completely.

The check is chosen by `lanczos(completeness = )` or
`options(eigencore.target_completeness = )`: `"auto"` (the default) counts edge targets
when the matrix is explicit and the predicted factorisation time (from
CHOLMOD's symbolic analysis, or \(n^3/3\) flops for dense) is at most
`max(0.5 s, solve time)` (options `eigencore.completeness_inertia_seconds`
and `eigencore.completeness_inertia_ratio`), and probes otherwise;
`"inertia"` always counts when it can; `"probe"` always probes; `"none"`
switches the check off. `certificate$completeness$inertia_gate` records why
the count was skipped. Matrices whose factors fill catastrophically (large
random sparse graphs) are left to the probe; banded, grid and other
mesh-like sparse matrices factor cheaply and are counted.

#### Every eigenvalue in an interval

The same counts drive the `interval(a, b)` target: `eig_partial()` counts
the eigenvalues in the closed interval `[a, b]` first (so `k` can be left
out), solves for exactly that many, and certifies the set with the count.
For a sparse matrix the interval is split into slices by further counts
("spectrum slicing"); each slice is solved by shift-invert Lanczos at its
centre and the slices are merged.

```{r interval-certificate}
L1 <- Matrix::bandSparse(30, k = c(0, 1),
                         diagonals = list(rep(2, 30), rep(-1, 29)),
                         symmetric = TRUE)
L2 <- kronecker(L1, Matrix::Diagonal(30)) + kronecker(Matrix::Diagonal(30), L1)
op <- options(eigencore.interval_dense_limit = 0)  # show the sparse route
fit_iv <- eig_partial(L2, target = interval(1, 1.5))
options(op)
length(values(fit_iv))
fit_iv$certificate$target_completeness
fit_iv$interval[c("count", "slices", "factorizations")]
```

The interval is closed: an eigenvalue numerically on an end point (inside
the zero band of the factorisation there) counts as inside. With residual
bound \(\rho\) from Kahan's theorem, if all `m` returned values lie inside
\([a, b]\) by more than \(\rho\), they are provably the `m` eigenvalues of
the interval (`"inertia_verified"`). A value within \(\rho\) of an end point
is listed in `completeness$boundary_ambiguous`; one recount on the interval
widened by \(\rho\) settles it (the same count proves the set; a larger one
means an eigenvalue sits just outside within the residual bound and the
result is `"inertia_inconclusive"`). A returned count different from the
interval count is `"inertia_failed"` and `passed = FALSE`. Dense matrices
use LAPACK's `dsyevr` with a value range instead, which selects the
eigenvalues by the same kind of Sturm count on the tridiagonal form
(`"exact"`).

## Comparing eigencore certificates to RSpectra diagnostics

`RSpectra::eigs_sym()` returns `nconv` and `niter` but does not return
residuals, backward errors, or an orthogonality measure. The eigencore
shim (`eigencore::eigs_sym()`) returns the same RSpectra-shaped list with
two added fields:

```{r rspectra}
res <- eigs_sym(A, k = 5, which = "LA")
names(res)
res$certificate
```

Code already written against `RSpectra::eigs_sym()` ignores `certificate`
and `diagnostics` silently; new code can opt in to certified results
without changing call sites.

## Cheat sheet

| You see                             | What it means                                                            | What to do                                                |
|-------------------------------------|--------------------------------------------------------------------------|-----------------------------------------------------------|
| `passed = TRUE` | All checks required by this certificate type passed; the 2-norm scale was exact or a lower bound, so the backward error is not under-stated. | Use the values/vectors within the stated certificate scope. |
| `passed = FALSE`, `failed_indices` non-empty | Some pairs hit `maxit` before converging.                                  | Increase `maxit`, raise `tol`, or request a wider `k`.    |
| `passed = FALSE`, `norm_bound_type` is `two_norm_lower_bound`, residuals tiny | The norm lower bound may be weak (e.g. smallest targets of a matrix-free operator). | Supply `metadata$two_norm` if `||A||_2` is known, or use a built-in matrix class. |
| `max_orthogonality_loss` near `sqrt(eps)` but residuals tiny | Iterative drift; clustered eigenvalues at risk of duplicates.             | Increase `maxit`; check whether there are repeated eigenvalues. |
| `failed_indices` includes the first returned pairs | Some leading returned pairs exceed the tolerance. | Inspect `convergence_history`; increase `maxit` if the errors are still declining. |
| `passed = FALSE`, `residual_passed = TRUE`, `target_completeness = "inertia_failed"` | An eigenvalue count proves a more-preferred eigenvalue is missing from the returned set. | Use a block method with a block at least the multiplicity (`lanczos(block = )`), raise `eigencore.completeness_max_rounds`, or a dense solve. |
| `target_completeness = "inertia_inconclusive"` | The target edge cuts through a cluster (closer than the residual bound); every returned pair is still certified. | Request a `k` that does not split the cluster, or tighten `tol`. |
| `passed = FALSE`, `residual_passed = TRUE`, `target_completeness = "failed"` | Every pair is accurate, but a more-preferred eigenvalue (e.g. a missed copy of a repeated value) lies outside the returned set. | Use a block method with a block at least the multiplicity (`lanczos(block = )`), raise `eigencore.completeness_max_rounds`, or a dense solve. |
| `failed_indices` includes the last returned pairs | Some trailing returned pairs exceed the tolerance. | Inspect the spectral gap; a wider `k` may help when the target boundary cuts through a cluster. |

The certificate records which numerical checks were run, the scale used for
each backward error, and whether those checks passed. Use those fields together
with the solver plan and the requirements of your downstream analysis.

## Where to go next

- `vignette("sparse-pca")` shows certificates for centered and scaled sparse
  operators in context, including operators that carry no norm metadata.
- `vignette("eigencore")` and `vignette("generalized-eigenproblems")` cover
  the workflows that produce the certificates read here.
