---
title: "Get started with eigencore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with eigencore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

eigencore turns spectral computation into a five-step workflow:

1. **Build an operator** describing the action of `A`.
2. **Describe the problem** (eigen or SVD; metric `B`; spectral target).
3. **Plan a solver** and inspect what it will run.
4. **Solve.**
5. **Inspect the certificate** — the numerical evidence that the result is trustworthy.

This vignette walks through the workflow on three problem classes: a standard
Hermitian eigenproblem, a generalized SPD eigenproblem with a metric `B`, and
a partial SVD. It also shows the RSpectra-compatible shim for users migrating
from existing code.

```{r setup}
library(eigencore)
suppressPackageStartupMessages(library(Matrix))
```

## 1. Standard Hermitian eigenproblem

Build a small symmetric matrix and ask for the five largest eigenvalues.

```{r}
set.seed(1)
n <- 200
A <- crossprod(matrix(rnorm(n * n), n, n)) / n + diag(n)
```

eigencore wraps the matrix in a *block-native operator* the moment you hand it
to a problem constructor. You can do this explicitly:

```{r}
Aop <- as_operator(A)
Aop
```

The operator carries dimensions, structure tags, and a flag indicating whether
the underlying storage has a native kernel (in this case dense double — yes).

### Build the problem and inspect the plan

```{r}
P    <- eigen_problem(A, structure = hermitian(), target = largest())
plan <- plan_solver(P, k = 5)
plan
```

The plan tells you *exactly* which kernel will be invoked. There is no
silent dispatch.

### Solve and read the certificate

```{r}
fit <- solve(P, k = 5)
fit
```

```{r}
fit$certificate
```

A passing certificate means: every requested pair has

- residual `||A v - lambda v|| / s` below `tol`, where `s` is the labeled norm
  bound (here `frobenius_exact`),
- backward error below `tol`,
- inter-vector orthogonality loss below the orthogonality tolerance.

If any pair fails, the certificate's `failed_indices` slot tells you which.

```{r}
fit$values
```

## 2. Generalized SPD eigenproblem (`A v = lambda B v`)

Pass a metric `B` to `eigen_problem()` (or to `eig_partial()` via the `B`
argument).

```{r}
B <- diag(seq(1, 5, length.out = n))
fit_gen <- eig_partial(A, k = 5, target = largest(), B = B,
                       method = lobpcg(maxit = 200))
fit_gen
```

The certificate's residual is `||A v - lambda B v|| / (||A|| + |lambda| ||B||)`,
and orthogonality is measured in the `B`-inner product where appropriate.

## 3. Partial SVD

For rectangular problems use `svd_partial()`:

```{r}
M <- matrix(rnorm(400 * 50), 400, 50)
svd_fit <- svd_partial(M, rank = 5, target = largest())
svd_fit
```

The reported `method` again identifies the path — for very small or
near-square problems eigencore may use a dense LAPACK SVD fallback rather than
running its iterative Golub–Kahan kernel; either way the certificate covers
both `||A v - sigma u||` and `||A^T u - sigma v||`.

## 4. RSpectra-compatible workflow

If your existing code uses `RSpectra::eigs_sym()`, you can call eigencore in
the same shape — the return list extends RSpectra's by adding `certificate`
and `diagnostics`:

```{r}
res <- eigs_sym(A, k = 5, which = "LA")
str(res, max.level = 1)
```

```{r}
res$certificate
```

`eigs()`, `eigs_sym()`, and `svds()` accept the same `which` codes as
`RSpectra` — `"LM"`, `"SM"`, `"LA"`, `"SA"`, `"LR"`, `"SR"`, `"LI"`, `"SI"`,
and `"BE"`.

## Where to go next

- The package reference, organised by topic, lives in the
  [reference index](../reference/index.html).
- `?certificate` documents the certificate fields in detail.
- `?plan_solver` explains how operator structure, target, and method
  combine to choose a kernel.
- `linear_operator()` lets you wrap a matrix-free `apply` callback as a
  first-class operator; eigencore's planner will warn when an R-level
  callback is being driven from a block-native solver hot loop, so you
  can decide whether to invest in a native operator implementation.
