---
title: "Migrating from geigen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from geigen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(eigencore)
library(Matrix)
```

eigencore is not a namespace clone of `geigen`. The migration target is the
problem class: dense SPD/Hermitian metrics, dense general pencils, dense QZ, or
sparse SPD partial solves. Results carry planner labels and certificates so
the numerical path is inspectable.

| geigen call | eigencore call | Status |
| --- | --- | --- |
| `geigen::geigen(A, B, symmetric = TRUE)` | `eig_full(A, B = ...)` or `eig_partial(A, B = ..., k = ...)` | Supported for SPD/Hermitian `B` |
| `geigen::geigen(A, B, symmetric = FALSE)` | `eig_full(A, B = ..., structure = general())` | Supported for dense general pencils |
| `geigen::gqz(A, B)` | `generalized_schur(A, B)` | Supported for dense QZ |
| `geigen::gevalues(qz)` | `values(qz)` and `alpha_beta(qz)` | Supported |
| `geigen::gsvd(A, B)` | `generalized_svd(A, B, ...)` | Deferred |

## Dense SPD/Hermitian Problems

For a full dense generalized SPD/Hermitian problem, pass `B` directly to
`eig_full()`.

```{r dense-spd}
A <- diag(c(2, 8, 18))
B <- diag(c(1, 2, 3))

fit <- eig_full(A, B = B)
values(fit)
fit$method
certificate(fit)$passed
```

For partial spectra, use `eig_partial()`.

```{r partial-spd}
part <- eig_partial(A, B = B, k = 2, target = smallest())
values(part)
part$method
certificate(part)$passed
```

## Dense General Pencils

Use `structure = general()` when `B` is indefinite, singular, nonsymmetric, or
when homogeneous `alpha`/`beta` diagnostics matter.

```{r dense-general}
A <- matrix(c(1, 4, 2, 3), 2, 2)
B <- matrix(c(2, 1, 0, -1), 2, 2)

pencil <- eig_full(A, B = B, structure = general())
values(pencil)
alpha_beta(pencil)$classification
pencil$method
```

Singular pencils expose finite, infinite, and undefined classifications rather
than hiding beta-zero cases.

```{r beta-zero}
singular <- eig_full(
  diag(c(2, 3, 0)),
  B = diag(c(1, 0, 0)),
  structure = general()
)

values(singular)
alpha_beta(singular)$classification
certificate(singular)$failed_indices
```

## QZ and gevalues

`geigen::gqz()` maps to `generalized_schur()`. Use `values()` for
`alpha / beta`, and `alpha_beta()` when the homogeneous coordinates or
classification are needed.

```{r qz}
qz <- generalized_schur(A, B)
values(qz)
coords <- alpha_beta(qz)
coords$alpha
coords$beta
coords$classification
qz$method
```

The current QZ sorting contract is eigencore-specific: use no sorting,
finite-first sorting, or infinite-first sorting. Other geigen sort modes are
not silently accepted.

```{r qz-sort}
beta_zero <- generalized_schur(
  diag(c(2, 3, 0)),
  diag(c(1, 0, 0)),
  sort = "infinite"
)
alpha_beta(beta_zero)$classification
```

## Sparse Users

Do not migrate sparse problems by relying on implicit densification. Full dense
surfaces require base dense matrices and reject sparse/operator inputs.

```{r sparse-reject, error = TRUE}
eig_full(Matrix::Diagonal(3), B = Matrix::Diagonal(3))
```

Sparse SPD/Hermitian generalized partial problems use `eig_partial()` with an
explicit no-densification policy when that boundary matters.

```{r sparse-partial}
A_sparse <- Matrix::Diagonal(x = c(1, 4, 9, 16, 25, 36))
B_sparse <- Matrix::Diagonal(x = c(1, 2, 3, 4, 5, 6))

sparse_fit <- eig_partial(
  A_sparse,
  B = B_sparse,
  k = 3,
  target = smallest(),
  method = lanczos(max_subspace = 6),
  allow_dense_fallback = "never"
)

values(sparse_fit)
sparse_fit$method
sparse_fit$restart$metric_solve
certificate(sparse_fit)$passed
```

## GSVD

`geigen::gsvd()` has no eigencore replacement yet. Keep using
`geigen::gsvd()` for GSVD workloads until `generalized_svd(A, B, ...)` is
implemented and documented.
