---
title: "Work with singular PSD geometry"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: structural
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Work with singular PSD geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    "});</script>"
  ),
  params$family,
  params$preset
))
```

```{r packages}
library(eigencore)
library(Matrix)
```

A positive-semidefinite matrix `K` can measure energy even when it is
singular. The singular directions have zero energy, so `K` is a seminorm on
the original coordinates and a genuine metric only after restricting to
`image(K)` or identifying points that differ by `null(K)`.

Eigencore 1.3 turns that geometry into a certified factor. You begin with four
facts:

1. the source shape and representation;
2. the real symmetric PSD form `K`;
3. the action you need, such as a form, root, projector, or pseudoinverse; and
4. the evidence the representation can supply.

The factor validates those facts once, records any repair, and fails before a
fallback when the requested action is not certified.

## Certify one singular dense form

This four-coordinate example has two positive directions and a two-dimensional
null space. The orthogonal rotation makes the null space nontrivial in the
original coordinate system.

```{r dense-factor}
Q <- matrix(c(
  1,  1,  1,  1,
  1, -1,  1, -1,
  1,  1, -1, -1,
  1, -1, -1,  1
), 4, 4, byrow = TRUE) / 2

K <- Q %*% diag(c(9, 1, 0, 0)) %*% t(Q)
K_factor <- psd_factor(K)
K_factor
c(rank = psd_rank(K_factor), nullity = psd_nullity(K_factor))
```

`psd_rank()` and the default `psd_nullity()` are numerical claims under the
factor's recorded tolerance. A generic floating-point eigendecomposition does
not prove an exact-arithmetic nullity theorem, so
`psd_nullity(K_factor, "algebraic")` is deliberately unavailable. Structural
constructors can make that stronger claim when their proof supports it.

Inspect the manifest before asking reusable code for an optional action:

```{r dense-capabilities}
psd_capabilities(K_factor)
certificate(K_factor)
```

The certificate distinguishes the source from the action that will run. Tiny
accepted negative eigenvalues, numerical-null positive eigenvalues, and
symmetry averaging are repairs, not erased history: their categories and
defects remain in `K_factor$spectrum` and `certificate(K_factor)`.

## Apply roots, projectors, and the pseudoinverse

All complete factors expose the same six square actions. `psd_apply()` accepts
a vector or a block with one row per original coordinate.

```{r dense-actions}
X <- matrix(c(
  1, 2, 3, 4,
  4, 3, 2, 1
), 4, 2)

KX <- psd_apply(K_factor, X, "form")
K_half_X <- psd_apply(K_factor, X, "sqrt")
image_X <- psd_apply(K_factor, X, "image_projector")
null_X <- psd_apply(K_factor, X, "null_projector")

stopifnot(
  isTRUE(all.equal(KX, K %*% X, tolerance = 1e-12)),
  isTRUE(all.equal(image_X + null_X, X, tolerance = 1e-12))
)
```

The inverse square root and pseudoinverse invert only retained positive modes.
They are not ordinary inverses of singular `K`:

```{r generalized-inverses}
K_half <- psd_apply(K_factor, diag(4), "sqrt")
K_ihalf <- psd_apply(K_factor, diag(4), "inverse_sqrt")
K_plus <- psd_apply(K_factor, diag(4), "pseudoinverse")
P <- psd_apply(K_factor, diag(4), "image_projector")

stopifnot(
  isTRUE(all.equal(K_half %*% K_half, K, tolerance = 1e-12)),
  isTRUE(all.equal(K_half %*% K_ihalf, P, tolerance = 1e-12)),
  isTRUE(all.equal(K %*% K_plus, P, tolerance = 1e-12))
)
```

Use `psd_operator(K_factor, action)` when an eigencore solver or another
operator consumer needs the action without a dense materialization.

## Move explicitly to the metric space

`psd_reduce()` maps original-coordinate columns into canonical image
coordinates. `psd_lift()` returns their minimum-Euclidean-norm representatives.
Null-space contamination disappears under reduction, while every K-Gram is
preserved.

```{r quotient}
reduced_X <- psd_reduce(K_factor, X)
contaminated_X <- X + 3 * null_X

stopifnot(
  isTRUE(all.equal(
    psd_reduce(K_factor, contaminated_X),
    reduced_X,
    tolerance = 1e-12
  )),
  isTRUE(all.equal(
    crossprod(reduced_X),
    psd_gram(K_factor, X),
    tolerance = 1e-12
  )),
  isTRUE(all.equal(
    psd_lift(K_factor, reduced_X),
    image_X,
    tolerance = 1e-12
  ))
)
```

This is the right boundary for algorithms that require an honest inner
product. `psd_orthonormalize()` performs the corresponding block operation and
reports discovered rank, conditioning, dropped directions, work, and a fresh
postcondition certificate.

```{r block}
K_basis <- psd_orthonormalize(K_factor, X, required_rank = 2)
K_basis
psd_gram(K_factor, K_basis$basis)
certificate(K_basis)$passed
```

## Distinguish pseudoinverse application from solving an equation

The Moore--Penrose action `K^+ b` is defined for every finite `b`; it silently
drops `b`'s null component. A strict solution of `K x = b` exists only when
`b` lies in `image(K)`. `psd_solve()` checks that compatibility per right-hand
side and certifies the original equation.

```{r strict-solve}
b <- c(1, 0, 0, 0)
psd_apply(K_factor, b, "pseudoinverse")

incompatible <- tryCatch(psd_solve(K_factor, b), error = identity)
c(
  class = class(incompatible)[1],
  code = incompatible$code,
  field = incompatible$field
)

b_image <- psd_apply(K_factor, b, "image_projector")
strict <- psd_solve(K_factor, b_image)
stopifnot(
  strict$compatible,
  certificate(strict)$passed,
  isTRUE(all.equal(
    as.numeric(K %*% strict$solution),
    as.numeric(b_image),
    tolerance = 1e-12
  ))
)
```

Relax the RHS tolerance only when that change is part of your model; the
factor records the default scale-relative policy, while an operation-specific
override is recorded in the solve certificate.

## Reduce a singular generalized eigenproblem

The existing generalized-eigen `B`/`metric=` surface remains SPD-only. Do not
pass a singular PSD form to it. Instead, reduce the problem explicitly to the
finite image space, solve the ordinary Euclidean problem there, then lift the
vectors.

```{r reduced-eigenproblem}
A <- Q %*% diag(c(8, 3, 5, 2)) %*% t(Q)
A_reduced <- psd_reduced_operator(K_factor, A)
A_image <- A_reduced$apply(diag(psd_rank(K_factor)))
fit <- eig_full(A_image)

largest_index <- which.max(values(fit))
value <- values(fit)[largest_index]
vector_original <- psd_lift(K_factor, vectors(fit)[, largest_index])
residual <- A %*% vector_original -
  value * psd_apply(K_factor, vector_original, "form")

c(value = value, residual_norm = sqrt(sum(residual^2)))
stopifnot(certificate(fit)$passed, sqrt(sum(residual^2)) < 1e-9)
```

Here the certified image has dimension two, so materializing its two-by-two
operator gives `eig_full()` an exact norm for its certificate. For a large
image, pass `A_reduced` to `eig_partial()` instead; an operator without exact
norm metadata can return accurate residuals while withholding an unqualified
`passed` flag because its certificate scale is estimated.

The caller still owns the scientific question: whether discarding the null
space preserves the requested estimand, and whether the retained rank is
adequate. Eigencore owns the numerical reduction, provenance, and certificate.

## Choose a constructor whose evidence matches the action

Identity and diagonal sources use analytic complete paths. Dense square
sources use a complete symmetric eigendecomposition. A supplied dense Gram
factor uses a complete compact SVD. These paths expose roots, projectors,
rank, reduction, strict solve, block primitives, and reduced operators.

```{r complete-constructors}
identity_factor <- psd_identity(4)
diagonal_factor <- psd_factor(c(9, 1, 0, 0))

L_dense <- matrix(c(
  1, 0,
  0, 1,
  1, 1
), 3, 2, byrow = TRUE)
gram_dense <- psd_gram_factor(L_dense, orientation = "columns")

c(
  identity = psd_rank(identity_factor),
  diagonal = psd_rank(diagonal_factor),
  dense_gram = psd_rank(gram_dense)
)
```

Sparse Gram and graph-Laplacian constructors preserve sparse state and make
only structural claims. A sparse Gram factor proves `K = L L^T` or
`K = L^T L`, so form and Gram actions are available without an `n` by `n`
dense matrix. It does not label `L` as the principal square root.

```{r sparse-gram}
L_sparse <- Matrix(L_dense, sparse = TRUE)
gram_sparse <- psd_gram_factor(L_sparse, orientation = "columns")
psd_capabilities(gram_sparse)
psd_apply(gram_sparse, diag(3), "form")
```

A sparse graph Laplacian proves positive semidefiniteness through its graph
structure and proves algebraic nullity by counting connected components.
Numerical rank, roots, projectors, and reduction remain unavailable because
the constructor never computes a complete spectrum.

```{r sparse-laplacian}
L_path <- bandSparse(
  5, 5,
  k = c(-1, 0, 1),
  diagonals = list(rep(-1, 4), c(1, 2, 2, 2, 1), rep(-1, 4))
)
laplacian <- psd_laplacian(L_path)

psd_nullity(laplacian, type = "algebraic")
psd_apply(laplacian, rep(1, 5), "form")

rank_error <- tryCatch(psd_rank(laplacian), error = identity)
c(class = class(rank_error)[1], code = rank_error$code)
```

The admitted release manifest is:

| Representation | Certified form/Gram | Root, pseudoinverse, projectors | Reduction, solve, block | Rank/nullity evidence | Dense `n` by `n` state |
|:---|:---:|:---:|:---:|:---|:---:|
| identity | yes | yes | yes | numerical and algebraic complete | no |
| diagonal | yes | yes | yes | numerical and algebraic complete | no |
| dense spectral | yes | yes | yes | numerical complete; algebraic unavailable | yes |
| dense Gram | yes | yes | yes | numerical complete; algebraic unavailable | yes |
| sparse Gram | yes | no | no | unavailable | no |
| sparse Laplacian | yes | no | no | algebraic component nullity only | no |
| generic sparse symmetric | rejected | no | no | unavailable | no |
| opaque callback | rejected without probing | no | no | unavailable | no reusable factor |

`psd_capabilities()` is the runtime authority. An unavailable cell raises a
typed `eigencore_psd_error` with a stable `code` and `capability`; it does not
densify or run an undocumented approximation.

## Understand tolerances, ownership, and persistence

All default thresholds use the admitted matrix's exact Frobenius scale:

```text
threshold = absolute tolerance + relative tolerance * ||K||_F.
```

No `max(1, ||K||_F)` floor is inserted, so zero-absolute-tolerance
classification is invariant under finite positive rescaling. Symmetry,
positivity, rank, and RHS compatibility have distinct typed tolerances because
they answer different questions. Customize them with `psd_policy()` and
`psd_tolerance()`, not unnamed scalar cutoffs.

Every 1.3 factor is an immutable snapshot. `operator_identity()`, `work()`, and
`retained_bytes()` expose lineage, logical work, and retained memory. Base
`saveRDS()`/`readRDS()` is the persistence format; integrity and schema checks
run before a restored factor acts. Live and opaque factors are not admitted in
1.3, even when a callback carries an operator ID and revision.

```{r opaque-failure}
callback_calls <- 0L
opaque <- linear_operator(
  dim = c(4, 4),
  apply = function(X, ...) {
    callback_calls <<- callback_calls + 1L
    X
  },
  dtype = "double",
  structure = hermitian(),
  name = "declared Hermitian callback"
)

opaque_error <- tryCatch(psd_factor(opaque), error = identity)
c(
  class = class(opaque_error)[1],
  code = opaque_error$code,
  callback_calls = callback_calls
)
```

This boundary keeps evidence separate from assertion: Hermitian metadata says
how an operator is intended to behave, but it is not a PSD proof, a complete
spectrum, or a certified square root.

## Downstream ownership

Application packages should consume exported factor actions and keep their
own estimands and algorithms:

- gprocrustes keeps translation gauges, O/SO choices, objectives, polar
  solvers, and consensus estimators;
- DKGE keeps design construction, scientific rank admission, cross-validation,
  regularization, contrasts, inference, and interpretation; and
- optimal-transport packages keep mass policy, Sinkhorn state, objectives, and
  convergence semantics.

An eigencore certificate establishes the numerical statement it names. It does
not decide that a singular geometry is scientifically adequate for a caller's
question.
