---
title: "Freeze and replay a solver decision"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Freeze and replay a solver decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Sometimes you need more than the answer to an eigenproblem: you need a record
of the solver decision that produced it. A problem describes what to compute,
whereas an executable plan freezes how eigencore will compute it. Saving the
plan lets you inspect, run, and replay the same route without asking the
planner to decide again.

This vignette follows one small Hermitian problem from construction through
execution and RDS persistence. It also shows how operator identity prevents a
saved workflow from quietly drifting to different data.

```{r setup}
library(eigencore)
```

## What does a plan add to a problem?

Start with a matrix and the numerical question: its two largest eigenpairs.
The problem records that question. `plan_solver()` then chooses a route and
freezes the controls, tolerance, operator identity, and planner policy used in
that choice.

```{r make-plan}
A <- diag(seq(200, 1))
problem <- eigen_problem(A, target = largest())
plan <- plan_solver(problem, k = 2, tol = 1e-10)
plan
```

Keep the distinction simple:

| Object | What it preserves | What happens when you solve it |
|:---|:---|:---|
| problem | operator, structure, and target | eigencore plans again under current policy |
| plan | problem plus a frozen route and execution contract | eigencore validates and runs that route |
| result | values, vectors, certificate, work, and governing plan | nothing is recomputed until you ask |

## How do you run and interpret the frozen decision?

`solve(plan)` validates the plan before applying the operator. The result keeps
both the planned route and the route that actually returned the answer. They
differ only when an allowed runtime fallback is needed.

```{r solve-plan}
fit <- solve(plan)
data.frame(
  planned = fit$planned_method,
  actual = fit$actual_method,
  fallback = fit$fallback_used,
  certified = certificate(fit)$passed
)
```

```{r solve-plan-checks, include = FALSE}
stopifnot(
  isTRUE(certificate(fit)$passed),
  isTRUE(all.equal(values(fit), c(200, 199), tolerance = 1e-10)),
  identical(fit$plan, plan),
  identical(fit$fallback_used, !identical(fit$planned_method, fit$actual_method))
)
```

If `fallback` is `TRUE`, inspect `fit$fallback_reason`. It names the failed
attempt, the route that returned the result, and a typed reason such as a
failed certificate. The certificate always describes the returned result, not
the abandoned attempt.

The work record uses logical units that can be compared without pretending
that every solver's historical `matvecs` field means the same thing:

```{r inspect-work}
w <- work(fit)
data.frame(
  complete = w$complete,
  solve_columns = w$operator_columns,
  certificate_columns = w$certification_operator_columns,
  iterations = w$iterations,
  restarts = w$restarts
)
```

`complete = TRUE` means every logical counter is known. When it is `FALSE`, an
`NA` counter means unknown, not zero; compare only the phases that both routes
report.

## How does the plan identify its operator?

Matrix-backed operators receive deterministic identities derived from their
dimensions, structure, and values. The revision therefore changes when the
matrix changes.

```{r inspect-identity}
identity_record <- operator_identity(plan)$A
data.frame(
  operator_id = identity_record$operator_id,
  revision = identity_record$revision,
  origin = identity_record$origin,
  portable = identity_record$portable
)
```

For a callback, eigencore cannot infer whether two functions represent the
same external data. Supply provenance only when your application can maintain
that promise:

```{r callback-identity}
callback <- linear_operator(
  dim = dim(A),
  apply = function(X, alpha = 1, beta = 0, Y = NULL) {
    Z <- alpha * (A %*% X)
    if (is.null(Y) || beta == 0) Z else Z + beta * Y
  },
  structure = hermitian(),
  metadata = list(frobenius_norm = sqrt(sum(A^2))),
  operator_id = "diagonal-example",
  revision = "values-v1",
  portable = TRUE
)
operator_identity(callback)[c("operator_id", "revision", "portable")]
```

`operator_id` names the logical lineage; `revision` names its current values
and behavior. Marking a callback portable is an assertion by its owner, not a
numerical test performed by eigencore. Without explicit provenance, a callback
receives an opaque session-local identity and its restored plan cannot run in a
different R session.

## How do you save and replay a plan?

Plans use ordinary RDS persistence. A restored portable plan validates its
schema, route, controls, execution arguments, policy snapshot, and operator
identity before the operator is called.

```{r rds-round-trip}
plan_path <- tempfile(fileext = ".rds")
saveRDS(plan, plan_path)
restored_plan <- readRDS(plan_path)
restored_fit <- solve(restored_plan)

data.frame(
  portable = restored_plan$serialization$portable,
  same_route = identical(restored_fit$planned_method, fit$planned_method),
  same_values = isTRUE(all.equal(values(restored_fit), values(fit)))
)
```

```{r rds-round-trip-checks, include = FALSE}
stopifnot(
  restored_plan$serialization$portable,
  identical(operator_identity(restored_plan), operator_identity(plan)),
  isTRUE(all.equal(values(restored_fit), values(fit), tolerance = 1e-10))
)
unlink(plan_path)
```

Replay means the same validated route and execution contract, not necessarily
bit-for-bit equality from an iterative method. Compare numerical answers under
the requested tolerance and inspect the fresh certificate on every result.

## What changes require a new plan?

Execution arguments are part of the frozen contract. An attempted override
fails with a typed error instead of producing a result under a route that no
longer matches the inspected plan.

```{r override-error, include = FALSE}
override_error <- tryCatch(
  solve(plan, tol = 1e-4),
  error = identity
)
stopifnot(
  inherits(override_error, "eigencore_plan_error"),
  identical(override_error$code, "execution_override"),
  identical(override_error$field, "tol")
)
```

```{r override-summary}
data.frame(
  class = class(override_error)[1],
  code = override_error$code,
  field = override_error$field
)
```

Use `solve(plan, replan = TRUE)` only when you deliberately want eigencore to
make a new decision from the embedded problem under current policy. The
original plan remains unchanged. If instead you want to reuse a certified
basis across related operators, use a restart state; that is a different
contract because the destination operator must receive a fresh certificate.

## Where should you go next?

- `vignette("warm-start-continuation")` shows basis reuse across related
  operators and the exact-revision versus changed-revision transition.
- `vignette("certificates")` explains the evidence carried by each result.
- `vignette("sparse-pca")` shows how planning exposes a non-densifying sparse
  route before computation begins.
- `?plan_solver`, `?operator_identity`, and `?work` document the complete
  records.
