---
title: PCA on sparse data without densifying
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{PCA on sparse data without densifying}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.1,
  fig.align = "center",
  out.width = "92%",
  dpi = 100
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r helpers, include = FALSE}
# Shared plotting helper. Kept out of the reader's view: the reader sees
# the API call and the picture, never the plotting boilerplate.
ec_blue <- "#2166ac"; ec_grey <- "grey78"

# A computed slice (blue) drawn on top of the full dense spectrum (grey).
plot_spectrum <- function(all_vals, computed, ylab = "value", main = NULL) {
  s <- sort(all_vals, decreasing = TRUE)
  k <- length(computed)
  op <- par(mar = c(4, 4.6, if (is.null(main)) 1 else 2.4, 1)); on.exit(par(op))
  plot(seq_along(s), s, pch = 19, cex = 0.4, col = ec_grey, bty = "n",
       xlab = "rank (largest to smallest)", ylab = ylab, main = main)
  points(seq_len(k), sort(computed, decreasing = TRUE),
         pch = 19, cex = 1.2, col = ec_blue)
  legend("topright", c("full spectrum (dense reference)", "computed by eigencore"),
         pch = 19, pt.cex = c(0.7, 1.2), col = c(ec_grey, ec_blue),
         bty = "n", cex = 0.9)
}
```

Principal component analysis starts with centering: subtract each column's
mean before finding directions of maximum variance. For a sparse matrix,
that one subtraction is the problem. `A - 1 %*% t(col_means)` fills in every
zero with a small nonzero number, so the matrix that started sparse becomes
dense before you've computed a single eigenvector.

eigencore solves the *centered* (and optionally scaled) problem as an
operator, so the subtraction never happens as a stored matrix. This
vignette builds one sparse dataset, centers and scales it, and computes a
certified partial SVD without ever forming the dense copy.

```{r setup}
library(eigencore)
library(Matrix)
```

## Build a sparse dataset with a low-rank signal

Simulate the kind of matrix behind collaborative filtering and topic
models: sparse interaction counts between 2,000 users and 500 items, built
from five latent factors so a handful of singular vectors should recover
most of the structure.

```{r simulate-signal}
set.seed(1)
m <- 2000L; n <- 500L; k_true <- 5L
U <- qr.Q(qr(matrix(rnorm(m * k_true), m, k_true)))
V <- qr.Q(qr(matrix(rnorm(n * k_true), n, k_true)))
signal <- U %*% diag(seq(8, 4, length.out = k_true)) %*% t(V)
```

Only 4% of user-item pairs are observed, and the observations are noisy:

```{r simulate-sparsify}
mask  <- rsparsematrix(m, n, density = 0.04)
noise <- rsparsematrix(m, n, density = 0.01, rand.x = function(k) rnorm(k, sd = 0.05))
A <- as((signal * (mask != 0)) + noise, "dgCMatrix")
dim(A)
```

```{r simulate-memory, echo = FALSE}
dense_mb  <- as.numeric(object.size(matrix(0, m, n))) / 1024^2
sparse_mb <- as.numeric(object.size(A)) / 1024^2
```

At 2,000 x 500 this is a small matrix, but the storage gap is already
visible: a dense copy would need `r sprintf("%.1f", dense_mb)` MB, while the
sparse representation needs `r sprintf("%.2f", sparse_mb)` MB — about
`r round(dense_mb / sparse_mb)` times less.

## Center without densifying

`center()` wraps `A` in an operator that subtracts column means on the fly,
computing the means directly from the sparse matrix rather than requiring
you to supply them.

```{r center}
Ac <- center(A, columns = TRUE)
Ac
```

Pass the centered operator straight to `svd_partial()`, exactly as you
would the original matrix.

```{r first-svd}
fit <- svd_partial(Ac, rank = 5, target = largest())
fit
```

```{r first-scree, echo = FALSE, fig.cap = "The five largest singular values of the centered matrix (blue) against the full singular spectrum (grey).", fig.alt = "Scatter plot of all 500 singular values sorted descending in grey, with the top five highlighted in blue."}
dense_centered <- scale(as.matrix(A), center = TRUE, scale = FALSE)
all_sv <- svd(dense_centered, nu = 0, nv = 0)$d
plot_spectrum(all_sv, fit$d, ylab = "singular value")
```

The five leading singular values separate cleanly from the rest of the
spectrum — exactly what you'd expect from data built around five latent
factors.

## Read the certificate on a composed operator

`fit$certificate$passed` is `FALSE` here, but look at why before treating
that as a problem:

```{r composed-certificate}
fit$certificate$max_backward_error
fit$certificate$norm_bound_type
fit$certificate$scale_is_estimate
```

The residual is tiny — far below the `1e-8` tolerance. What's withheld is
the *scale*: computing an exact Frobenius norm bound for a centered operator
would mean a second full pass over the data, so eigencore instead reports a
stochastic (Hutchinson) estimate and refuses to mark the certificate
`passed` while that estimate is the only evidence. This is the same honest
bookkeeping `vignette("certificates")` covers in detail — a composed
operator trades the exact bound for an estimated one unless you supply
norm metadata yourself, which the next section shows.

## Does it match what you'd get by densifying?

The whole point of `center()` is that it should be numerically
indistinguishable from centering the dense matrix directly. Check it:

```{r verify-dense}
max(abs(sort(fit$d, decreasing = TRUE) - sort(all_sv[1:5], decreasing = TRUE)))
```

The two agree to machine precision. `dense_centered` above was built only
to draw the comparison plot and run this check — the eigencore computation
itself never materialized it.

## Scale columns too

A common next step in PCA preprocessing is scaling each column to unit
variance, turning a covariance-matrix PCA into a correlation-matrix one.
Compute the column variances from the sparse matrix and pass them to
`scale_cols()`, composed with the centering operator you already have:

```{r scale-cols}
col_var <- Matrix::colMeans(A^2) - Matrix::colMeans(A)^2
w <- 1 / sqrt(col_var + 1e-6)
Acs <- scale_cols(Ac, w)
fit_scaled <- svd_partial(Acs, rank = 5, target = largest())
fit_scaled$d
```

`center()` and `scale_cols()` compose freely because both return the same
`eigencore_operator` type that every solver accepts — there's no separate
"scaled matrix" object to keep track of.

## What did the planner actually do?

Every solve is preceded by a plan, and you can inspect it directly instead
of trusting a black box:

```{r plan}
plan <- plan_solver(svd_problem(Acs), rank = 5, target = largest())
plan$method
plan$reasons
```

Centering and scaling a sparse matrix still fuse into a single native,
non-densifying construction. But once that composed operator reaches the
solver, eigencore drives it through a matrix-free Golub-Kahan callback
cycle rather than a fully block-native kernel — the `reasons` say so
explicitly rather than leaving you to infer it from timing. `plan_solver()`
takes the same problem/rank/target arguments as the solve itself, so you
can always check the plan before paying for the computation.

## Going fully matrix-free

Sometimes there's no matrix at all — just a function that knows how to
apply `A` and its adjoint. `linear_operator()` wraps arbitrary callbacks the
same way `center()` wraps a sparse matrix.

```{r matrix-free}
op <- linear_operator(
  dim = dim(A),
  apply = function(X, alpha = 1, beta = 0, Y = NULL) {
    # Sparse %*% dense returns an S4 dgeMatrix; as.matrix() keeps the
    # callback's return type consistent with what the native solver expects.
    Z <- alpha * as.matrix(A %*% X)
    if (is.null(Y) || beta == 0) Z else Z + beta * Y
  },
  apply_adjoint = function(X, alpha = 1, beta = 0, Y = NULL) {
    Z <- alpha * as.matrix(crossprod(A, X))
    if (is.null(Y) || beta == 0) Z else Z + beta * Y
  },
  name = "matrix-free A"
)
fit_mf <- svd_partial(op, rank = 5, target = largest())
fit_mf$certificate$norm_bound_type
```

Without any norm metadata, this is another stochastic-estimate certificate.
Supply the exact Frobenius norm — cheap to compute once for a sparse
matrix — and the certificate stops withholding `passed`:

```{r matrix-free-hint}
op_hinted <- linear_operator(
  dim = dim(A),
  apply = op$apply,
  apply_adjoint = op$apply_adjoint,
  name = "matrix-free A (exact norm)",
  metadata = list(frobenius_norm = norm(A, type = "F"))
)
fit_hinted <- svd_partial(op_hinted, rank = 5, target = largest())
fit_hinted$certificate$norm_bound_type
fit_hinted$certificate$passed
```

Whether you reach for `center()`/`scale_cols()` or hand-write a
`linear_operator()`, the certificate always tells you which kind of
evidence backed the result.

## Why this matters at scale

The 2,000 x 500 example above keeps this vignette fast to build, but the
memory argument only gets sharper as the matrix grows. A production
recommender with 5 million users and 200,000 items would need roughly 8 TB
to store as a dense matrix; the sparse representation, at typical
interaction densities, fits in a few gigabytes. Densifying to center it
would defeat the entire point of using a sparse format.

## Where to go next

- `vignette("certificates")` — the deep dive on `passed`, `scale_is_estimate`,
  and what to do about a withheld certificate.
- `vignette("eigencore")` — the general get-started workflow: operators,
  problems, plans, and solves for both eigenproblems and SVD.
- `vignette("generalized-eigenproblems")` — problems of the form
  `A v = lambda B v`, including whitened and metric-weighted variants.
- `?linear_operator` and `?compose` document the full operator algebra,
  including `crossprod_operator()` for building `A^* A` when an
  eigenproblem is a more natural fit than an SVD.
