---
title: "Reuse an eigenspace across a parameter sweep"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Reuse an eigenspace across a parameter sweep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

When you solve a sequence of nearby Hermitian eigenproblems, the eigenvectors
from one solve are often good starting directions for the next. Passing those
directions through `initial_subspace` can reduce operator work while preserving
the ordinary result contract: eigencore still recomputes the Ritz values,
residuals, orthogonality checks, and certificate for the current operator.

This vignette follows the smallest eigenpairs of `A - rho B` as `rho` changes.
At the end you will have a certified result at every step, provenance showing
whether the supplied start was used, and operator-column counts for comparing
the warm and cold solves.

```{r setup}
library(eigencore)
```

## What problem are we solving?

Here `B` is a perturbation term in the standard eigenproblem, not the optional
generalized-eigenproblem metric argument of `eig_partial()`. Both `A` and `B`
are symmetric, so every member of the family `A - rho B` is Hermitian.

```{r build-family}
n <- 180L
k <- 5L
A <- diag(2, n)
A[row(A) == col(A) + 1L] <- -1
A[row(A) + 1L == col(A)] <- -1
x <- seq_len(n) / (n + 1)
B <- diag(0.5 + x + 0.2 * sin(4 * pi * x))
```

Use an explicit Lanczos method for a warm start. This prevents `auto()` from
routing a structured or nearest-target problem to shift-invert, which does not
consume `initial_subspace`.

```{r first-solve}
method <- lanczos(
  block = k,
  max_subspace = 8L * k,
  max_restarts = 200L
)
first <- eig_partial(
  A, k = k, target = smallest(), method = method,
  tol = 1e-8, seed = 2026
)
certificate(first)$passed
```

## How do you reuse the previous eigenspace?

Pass the returned eigenvectors to the next solve. The values alone are not
enough: `initial_subspace` expects directions with one row per operator
dimension.

```{r warm-step}
rho <- 0.02
warm <- eig_partial(
  A - rho * B, k = k, target = smallest(), method = method,
  tol = 1e-8, seed = 2026, initial_subspace = vectors(first)
)
certificate(warm)$passed
```

The same pattern extends to a sweep. This example also runs a cold solve at
each step so the work and answers can be compared on equal terms.

```{r continuation-sweep}
rhos <- c(0.02, 0.04, 0.06)
start <- vectors(first)
rows <- vector("list", length(rhos))

for (i in seq_along(rhos)) {
  op <- A - rhos[[i]] * B
  cold <- eig_partial(
    op, k, smallest(), method = method, tol = 1e-8, seed = 2026
  )
  warm <- eig_partial(
    op, k, smallest(), method = method, tol = 1e-8, seed = 2026,
    initial_subspace = start
  )
  rows[[i]] <- data.frame(
    rho = rhos[[i]],
    cold_columns = cold$operator_columns,
    warm_columns = warm$operator_columns,
    max_value_difference = max(abs(sort(values(cold)) - sort(values(warm))))
  )
  start <- vectors(warm)
}
comparison <- do.call(rbind, rows)
```

```{r continuation-checks, include = FALSE}
stopifnot(
  isTRUE(certificate(first)$passed),
  all(vapply(rows, function(x) all(is.finite(unlist(x))), logical(1))),
  all(comparison$max_value_difference < 1e-8),
  all(comparison$warm_columns < comparison$cold_columns),
  isTRUE(certificate(warm)$passed)
)
```

```{r continuation-table, echo = FALSE}
shown <- comparison
shown$max_value_difference <- format(
  shown$max_value_difference, digits = 3, scientific = TRUE
)
knitr::kable(
  shown,
  col.names = c(
    "rho", "cold operator columns", "warm operator columns",
    "max |cold value - warm value|"
  )
)
```

On this reproducible family, every continuation step uses fewer operator
columns than its cold counterpart and agrees on the requested values well
within the solve tolerance. These counts include the work used to guard and
certify the supplied subspace; `operator_block_calls` is a different metric
because one block call may apply the operator to several columns.

Do not assume the same speedup for every sweep. When the target eigenspace
changes abruptly or the supplied directions have little overlap with it, a
warm solve can cost about as much as a cold one. The reproducible benchmark
`inst/benchmarks/bench-warm-start-continuation.R` includes both high-overlap
continuation and a deliberate overlap-loss jump.

## How can you tell what happened to the start?

`diagnostics()` exposes the start source and the boundary-processing counts.

```{r provenance}
d <- diagnostics(warm)
data.frame(
  start_source = d$start_source,
  supplied = d$initial_subspace$supplied,
  accepted = d$initial_subspace$accepted,
  rejected = d$initial_subspace$rejected,
  augmented = d$initial_subspace$augmented,
  rank = d$initial_subspace$rank
)
```

The supplied matrix must be numeric, finite, and have `n` rows. Eigencore
orthonormalizes its columns and detects numerical rank at the solver boundary.
If too few independent directions survive, the start block is augmented; if
more directions survive than the method's block width, a seeded rotation lets
all accepted directions contribute to the fitted block. The `seed` therefore
makes augmentation and compression reproducible as well as controlling a cold
random start.

Passing `initial_subspace = NULL` is exactly the cold-start contract. It leaves
the previous random-start sequence and result unchanged.

## Which solver plans accept a warm start?

The supported surface is deliberately narrow:

| Problem and plan | Warm-start status |
|---|---|
| Standard real Hermitian Lanczos, explicit dense double matrix | Native and supported |
| Standard real Hermitian Lanczos, `dgCMatrix` | Native and supported |
| Standard real Hermitian Lanczos, matrix-free operator, `block > 1` | Native callback path and supported |
| Standard real Hermitian Lanczos, matrix-free operator, `block = 1` | Reference-labelled path and supported |
| Generalized problem using the `B` argument | Rejected |
| Shift-invert or `nearest()` plan | Rejected |
| Dense fallback, nonsymmetric, or non-Lanczos plan | Rejected |

Use `method = lanczos()` when the start is required. With `method = auto()`,
inspect `plan_solver()` first: sparse algebraic-edge targets and
`nearest()` may select a factorized shift-invert plan. Supplying a start to an
unsupported plan raises an error instead of silently ignoring it.

For native block paths, `lanczos(check_stride = N)` with a positive integer
checks convergence every `N` block iterations and can stop a strong warm start
before a full cold-sized sweep completes. The default `check_stride = 0L`
preserves full-sweep behavior. Mid-sweep checks do not add operator
applications; enable them when workload-specific measurements show a benefit.

## What does the fresh certificate prove?

A starting subspace is only a hint. Each solve certifies the returned
eigenpairs against the current `A - rho B`; it never reuses the previous
certificate.

There is one important distinction: a small residual proves that a returned
pair is an eigenpair, but not that it belongs to the requested end of the
spectrum. A fully supplied subspace that is already invariant at the requested
tolerance is therefore discarded in favor of a cold start, and
`start_source` records the invariant-guard decision. For reliable continuation,
supply directions that overlap the requested target eigenspace and always
check `certificate(fit)$passed`.

## Where should you go next?

- `?eig_partial` documents the complete `initial_subspace` contract.
- `vignette("certificates")` explains residuals, backward error, and
  orthogonality checks.
- `diagnostics(fit)` reports start provenance and separates operator block
  calls, operator columns, and certification columns.
- Run `Rscript inst/benchmarks/bench-warm-start-continuation.R --strict` for
  the non-quick continuation evidence on your own machine.
