---
title: "Benchmarks: what we measure and what it means"
output:
  rmarkdown::html_document:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
---

This is a pkgdown-only article, not a CRAN vignette. It performs **no timing
at build time**: every number below is read from result directories stored in
the repository under `inst/benchmarks/results/`, produced by the benchmark
suite `inst/benchmarks/run-suite.R`. Rebuilding the site never changes a
number; adding a result directory does.

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.width = 8,
  fig.height = 4.4,
  fig.align = "center",
  out.width = "96%",
  dpi = 100
)
options(knitr.kable.NA = "-")
`%||%` <- function(x, y) if (is.null(x) || length(x) == 0L) y else x
html_marks <- c("&dagger;", "&Dagger;")  # locale-independent flags in tables

find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, mustWork = FALSE)
  for (i in 1:6) {
    if (file.exists(file.path(d, "DESCRIPTION")) &&
        file.exists(file.path(d, "inst", "benchmarks", "report.R"))) return(d)
    d <- dirname(d)
  }
  NULL
}
repo_root <- find_repo_root()
if (is.null(repo_root)) repo_root <- find_repo_root(knitr::current_input(dir = TRUE) %||% ".")
have_report <- !is.null(repo_root)
if (have_report) source(file.path(repo_root, "inst", "benchmarks", "report.R"))
results_root <- if (have_report) file.path(repo_root, "inst", "benchmarks", "results") else NULL
std_dirs <- if (have_report) bench_find_runs(results_root, profile = "standard") else character()
scal_dirs <- if (have_report) bench_find_runs(results_root, profile = "scaling") else character()
std <- if (length(std_dirs)) bench_load(std_dirs) else NULL
scal <- if (length(scal_dirs)) bench_load(scal_dirs) else NULL
have_std <- !is.null(std)
have_scal <- !is.null(scal)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r no-results, eval = !have_std, results = 'asis'}
cat(paste0(
  "> **No stored benchmark results were found.** This article renders from ",
  "`inst/benchmarks/results/<date>-<machine>-<profile>/`. Run the suite ",
  "(see *How to reproduce* below) and rebuild the site.\n"
))
```

## What is measured and why

Each **case** is a deterministic problem (fixed generator and seed) with a
wanted set: the `k` largest or smallest eigenvalues, the `k` eigenvalues of
largest modulus, the `k` eigenvalues nearest a shift, the `k` largest singular
triplets, or the full spectrum. Every **method** (eigencore, RSpectra, irlba,
PRIMME when installed, and dense base R/LAPACK where it fits in memory) is run
on the same matrix with the same requested tolerance (`tol = 1e-8`) and, for
each case and method, the suite records:

- **wall time**: one untimed warm-up call, then the median and minimum over
  `reps` timed calls (5 in the standard profile). eigencore's time always
  includes building its certificate; the other packages do no certification.
- **operator applications** (`matvecs`): the hardware-independent work each
  package reports about itself: eigencore's `work()` record
  (`operator_columns + adjoint_columns` of the solve phase), RSpectra's `nops`,
  irlba's `mprod`, PRIMME's `numMatvecs`. For shift-invert cases an
  application is a linear solve. eigencore's explicit-Gram SVD paths do not
  apply the operator column by column, so they report no comparable count.
- **memory**: peak resident-set growth of the process during the warm-up call
  (Linux, via `VmHWM`), which includes C/C++ heaps, and the R heap high-water
  mark. Both are in `results.csv`; neither is a cross-package claim.
- **accuracy**, computed by the suite itself for every method's output,
  defined in the next section, plus eigencore's own certificate verdict.

Threads are a property of the run: each thread setting runs in a fresh R
process with `OPENBLAS_NUM_THREADS`, `OMP_NUM_THREADS`, and
`options(eigencore.threads)` all set to the same value, so "4 threads" means
the whole process (BLAS, LAPACK and eigencore's OpenMP kernels) had four.
RSpectra and irlba are effectively single-threaded on sparse inputs, so their
4-thread rows mostly measure noise.

## The independent accuracy check

The suite does not trust any solver's own residuals. For every returned pair
it recomputes

- the residual `A x - lambda B x` (`B = I` for standard problems), and
  for SVDs both sides,
  `sqrt(||A v - sigma u||^2 + ||A' u - sigma v||^2)`;
- the **normwise 2-norm backward error**
  `||r|| / ((||A||_2 + |lambda| ||B||_2) ||x||)` (SVD: `/ ||A||_2`). This is
  the definition eigencore's own certificates use since the C12 change, but
  the suite divides by a *high-accuracy* `||A||_2`: exact when the spectrum is
  known or a dense factorisation is affordable, otherwise a converged
  Golub-Kahan-Lanczos estimate with full reorthogonalisation, computed once
  per case and cached. (eigencore's certificate divides by a value that never
  exceeds `||A||_2`, so its own number can only be larger.)
- the orthogonality loss `max |X' X - I|` (`X' B X - I` for generalized
  problems, `U` and `V` for SVDs);
- the **value error** against a trusted reference: the analytic spectrum
  (Laplacians, finite-element pencils, matrices built as `Q diag(values) Q'`),
  dense LAPACK when feasible, or else a tight iterative reference: RSpectra
  at `tol = 1e-13` asking for `k + 10` and `k + 20` values with large
  subspaces, and eigencore at `tol = 1e-12`. Each candidate is verified with
  the independent backward error, and among the accurate ones the set
  reaching furthest into the target wins (a value with a tiny backward error
  is a genuine eigenvalue, so a set that reaches further cannot be wrong where
  a shorter one may have skipped values). Disagreement between candidates is
  stored in `cases.csv`. The error is scaled like the
  backward error: `|lambda - lambda_ref| / (||A||_2 + |lambda_ref| ||B||_2)`.
- whether the method returned the **requested target set** (`target_ok`):
  `k` values, each matched to a distinct reference value within a gap-aware
  tolerance, and every matched reference value inside the wanted set (ties at
  the boundary, such as repeated eigenvalues and conjugate pairs, count as
  inside). A small backward error does not imply the right target set: a
  Krylov method can converge accurately to the wrong eigenvalues, or return
  one copy of a repeated eigenvalue and skip the other.

In the tables below a time is marked † when the method did **not** return
the wanted set, and ‡ when its independent backward error exceeds `1e-6`.
Such times are not comparable with correct ones.

## Result sets

```{r env-table, eval = have_std}
env_tab <- bench_env_table(std)
if (have_scal) env_tab <- rbind(env_tab, bench_env_table(scal))
knitr::kable(t(env_tab), col.names = env_tab$run,
             caption = "Environment of each stored result set (one column per run).")
```

```{r case-table, eval = have_std}
cs <- std$cases[!duplicated(std$cases$case_id),
                c("case_id", "description", "nnz", "reference_source", "norm2_source")]
cs$nnz <- format(cs$nnz, big.mark = ",")
knitr::kable(cs, row.names = FALSE,
             col.names = c("case", "generator", "nnz", "reference values", "||A||_2"),
             caption = "Cases of the standard profile.")
```

## Standard profile results

```{r family-tables, eval = have_std, results = 'asis'}
r <- std$results
fam_names <- c(sym_sparse = "Sparse symmetric, extreme eigenvalues",
               nonsym_sparse = "Sparse nonsymmetric, largest modulus",
               svd_sparse = "Sparse SVD",
               dense_sym = "Dense symmetric, partial",
               eig_full = "Full dense eigendecomposition",
               shift_invert = "Interior eigenvalues by shift-invert",
               generalized = "Generalized symmetric-definite",
               pca = "Centred sparse PCA (implicit centring)",
               banded = "Banded (1-D Laplacian), smallest",
               laplacian2d = "2-D Laplacian, smallest (repeated eigenvalues)",
               powerlaw = "Power-law spectrum",
               clustered = "Clustered / repeated top eigenvalues",
               lowrank = "Low rank plus noise SVD",
               suitesparse = "SuiteSparse Matrix Collection")
cell <- function(x, f) if (!nrow(x)) "-" else f(x[1L, ])
tfmt <- function(x) if (x$status == "skipped") "n/a" else if (x$status != "ok") "error" else
  paste0(bench_fmt_time(x$time_median), bench_flag(x, html_marks))
for (fam in intersect(names(fam_names), unique(r$family))) {
  cat("\n### ", fam_names[[fam]], "\n\n", sep = "")
  rf <- r[r$family == fam, , drop = FALSE]
  keys <- unique(rf[, c("case_id", "threads")])
  keys <- keys[order(match(keys$case_id, unique(rf$case_id)), keys$threads), ]
  tab <- do.call(rbind, lapply(seq_len(nrow(keys)), function(i) {
    x <- rf[rf$case_id == keys$case_id[i] & rf$threads == keys$threads[i], , drop = FALSE]
    g <- function(m) x[x$method == m, , drop = FALSE]
    ec <- g("eigencore"); rs <- g("RSpectra")
    ratio <- if (nrow(ec) && nrow(rs) && ec$status == "ok" && rs$status == "ok")
      bench_fmt_ratio(ec$time_median / rs$time_median) else "-"
    mv <- function(z) if (!nrow(z) || z$status != "ok" || !is.finite(z$matvecs)) "-" else format(z$matvecs)
    acc <- function(z) if (!nrow(z) || z$status != "ok") "-" else
      sprintf("%s%s", bench_fmt_sci(z$max_backward_error),
              if (isFALSE(z$target_ok)) " (miss)" else "")
    data.frame(case = keys$case_id[i], threads = keys$threads[i],
               eigencore = cell(ec, tfmt), RSpectra = cell(rs, tfmt),
               irlba = cell(g("irlba"), tfmt), PRIMME = cell(g("PRIMME"), tfmt),
               base = cell(g("base"), tfmt), `eigencore / RSpectra` = ratio,
               `matvecs ec / RS` = paste(mv(ec), "/", mv(rs)),
               `bwd err ec` = acc(ec), `bwd err RS` = acc(rs),
               certified = if (nrow(ec) && !is.na(ec$eigencore_certified)) ifelse(ec$eigencore_certified, "yes", "NO") else "-",
               check.names = FALSE, stringsAsFactors = FALSE)
  }))
  drop <- vapply(tab, function(col) all(col %in% c("-", "- / -")), logical(1))
  print(knitr::kable(tab[, !drop, drop = FALSE], row.names = FALSE, align = "l"))
  cat("\n")
}
```

### Time relative to RSpectra

Ratios below one mean faster than RSpectra on the same machine and thread
setting.

```{r ratio-plot, eval = have_std, fig.height = 5.2}
rr <- std$results
ec <- rr[rr$method == "eigencore" & rr$status == "ok", c("case_id", "threads", "time_median", "target_ok")]
rs <- rr[rr$method == "RSpectra" & rr$status == "ok", c("case_id", "threads", "time_median", "target_ok")]
m <- merge(ec, rs, by = c("case_id", "threads"), suffixes = c("_ec", "_rs"))
if (nrow(m)) {
  m$ratio <- m$time_median_ec / m$time_median_rs
  ids <- unique(m$case_id)
  op <- par(mar = c(4.2, 17, 1, 1))
  y <- match(m$case_id, ids) + ifelse(m$threads == min(m$threads), -0.15, 0.15)
  plot(m$ratio, y, log = "x", yaxt = "n", ylab = "", bty = "n",
       xlab = "eigencore time / RSpectra time (log scale)",
       pch = ifelse(m$threads == min(m$threads), 19, 1), col = "#2166ac",
       ylim = c(0.5, length(ids) + 0.5))
  axis(2, at = seq_along(ids), labels = ids, las = 1, cex.axis = 0.75)
  abline(v = 1, lty = 2, col = "grey50")
  legend("bottomright", paste("threads", sort(unique(m$threads))), pch = c(19, 1),
         col = "#2166ac", bty = "n", cex = 0.85)
  par(op)
}
```

### Operator applications

Operator applications are the hardware-independent part of the comparison:
they do not change with the machine, the BLAS, or the load. A dash means the
package reported no comparable count (dense LAPACK, explicit Gram paths).

```{r matvec-table, eval = have_std}
mt <- bench_matvec_table(std, threads = min(std$results$threads))
knitr::kable(mt[, setdiff(names(mt), c("run_id", "threads"))], row.names = FALSE,
             caption = "Operator applications (A and A' counted separately) at threads = 1.")
```

### Accuracy summary

```{r accuracy-table, eval = have_std}
at <- bench_accuracy_table(std, threads = min(std$results$threads))
knitr::kable(at[, setdiff(names(at), c("run_id", "threads"))], row.names = FALSE,
             caption = paste("Independent backward error / scaled value error / target set",
                             "(\"cert\": eigencore's own certificate passed).",
                             "threads = 1."))
```

```{r findings, eval = have_std, results = 'asis'}
rr <- std$results
missed <- !is.na(rr$target_ok) & !rr$target_ok
cert_failed <- !is.na(rr$eigencore_certified) & !rr$eigencore_certified
bad <- rr[rr$status == "ok" & (missed | cert_failed), ]
bad <- bad[!duplicated(bad[, c("case_id", "method")]), ]
if (nrow(bad)) {
  cat("**Rows where a method missed the wanted set or eigencore's certificate failed** ",
      "(generated from the stored results):\n\n", sep = "")
  for (i in seq_len(nrow(bad))) {
    b <- bad[i, ]
    cat(sprintf("- `%s`, %s: %s%s (value error %s, backward error %s).\n",
                b$case_id, b$method,
                if (isFALSE(b$target_ok)) "returned the wrong or an incomplete target set" else "target set ok",
                if (isFALSE(b$eigencore_certified)) "; eigencore certificate **failed**" else
                  if (isTRUE(b$eigencore_certified) && isFALSE(b$target_ok)) "; eigencore's certificate nevertheless passed" else "",
                bench_fmt_sci(b$value_err), bench_fmt_sci(b$max_backward_error)))
  }
  cat("\nA residual certificate checks that each returned pair is accurate; it ",
      "does not by itself prove that no wanted eigenvalue was skipped, which is ",
      "what the target-set column measures.\n", sep = "")
}
```

## Scaling

```{r no-scaling, eval = !have_scal, results = 'asis'}
cat("> No stored `scaling` profile results were found.\n")
```

The scaling profile sweeps one variable at a time: the dimension `n` of a
random sparse symmetric matrix (5 nonzeros per row), its density at
`n = 20000`, the number of wanted eigenvalues `k`, the long side of a sparse
`m x 2000` SVD, and the dimension of the 1-D Laplacian, whose smallest
eigenvalues have gaps shrinking like `1/n^2`. Solid lines are 1 thread, dashed
lines 4 threads; open symbols mark results that missed the wanted set.

```{r scaling-plots, eval = have_scal, fig.height = 4}
groups <- intersect(c("scaling_n", "scaling_nnz", "scaling_k", "scaling_svd_m", "scaling_banded_n"),
                    unique(scal$results$group))
thr <- sort(unique(scal$results$threads))
for (g in groups) {
  op <- par(mfrow = c(1, 2))
  bench_plot_scaling(scal, g, threads = thr, metric = "time_median", main = paste(g, "- time"))
  bench_plot_scaling(scal, g, threads = min(thr), metric = "matvecs", main = paste(g, "- operator applications"))
  par(op)
}
```

```{r scaling-slopes, eval = have_scal}
sl <- do.call(rbind, lapply(groups, function(g) cbind(group = g, bench_scaling_slopes(scal, g))))
sl$slope <- round(sl$slope, 2)
knitr::kable(sl, row.names = FALSE,
             caption = "Fitted log-log slope of median time against the swept variable.")
```

## How to reproduce or add your machine

Install eigencore from the commit you want to measure into a private library
and run the suite from the repository root:

```sh
R CMD INSTALL --preclean --no-docs -l .rlib .
Rscript inst/benchmarks/run-suite.R --lib=.rlib --profile=quick              # ~2 min smoke run
Rscript inst/benchmarks/run-suite.R --lib=.rlib --profile=standard --threads=1,4
Rscript inst/benchmarks/run-suite.R --lib=.rlib --profile=scaling  --threads=1,4
```

Useful options: `--families=sym_sparse,svd_sparse` (or a sweep such as
`scaling_k`), `--cases=<substring>`, `--methods=eigencore,RSpectra,irlba,base,PRIMME`,
`--reps=N`, `--suitesparse` (adds a few SuiteSparse Matrix Collection matrices,
downloaded once into the cache and skipped when offline), and `--list` to
print the selected cases. Each run writes
`inst/benchmarks/results/<YYYYMMDD>-<machine-id>-<profile>/` containing
`results.csv` (one row per case, method and thread setting), `cases.csv`
(generators, reference sources and `||A||_2`), and `environment.json`/`.rds`
(CPU, cores, R, BLAS/LAPACK library and threads, package versions, eigencore
git SHA, load average). The machine id is a hash; no hostname is stored.

To add your machine, commit that directory; this article and the README
table pick up the newest run per machine and profile. Summaries are available
without rebuilding the site:

```sh
Rscript inst/benchmarks/report.R inst/benchmarks/results/<run-dir>
```

The `benchmarks` GitHub workflow (manual dispatch, inputs `profile` and
`threads`) runs the same command on `ubuntu-latest` with OpenBLAS and uploads
the result directory as an artifact; it never commits results.

## Limits

- **Shared, noisy hardware.** Stored results may come from a shared container
  whose load varies while the suite runs; the environment table records the
  load average at the start and end, and `results.csv` keeps every repetition
  (`times`) and the per-row load. Medians absorb short spikes, not sustained
  contention. Multithreaded rows are the most sensitive: an OpenMP region on
  an oversubscribed machine can be an order of magnitude slower than on an
  idle one, which is why matvec counts are reported alongside times.
- **BLAS dependence.** Dense cases and eigencore's blocked kernels depend on
  the BLAS/LAPACK library and its threads; RSpectra uses Eigen's own kernels.
  Compare rows only within one result set, and check the BLAS line of its
  environment.
- **Random matrices are not real matrices.** Most cases are synthetic with
  known structure (random sparsity, Laplacians, prescribed spectra). Real
  matrices have different sparsity patterns, spectra and conditioning; the
  optional SuiteSparse cases are a first step, and your own matrices are the
  only benchmark that settles a choice.
- **Different stopping rules.** All methods receive `tol = 1e-8`, but each
  package interprets it differently (RSpectra relative to the Ritz value,
  irlba relative to the largest singular value, eigencore as a backward error
  with a certificate). The independent accuracy columns are there so that a
  fast but inaccurate or incomplete answer is visible.
- **What the times include.** eigencore's time includes certification and
  result construction; dense base R rows include `as.matrix()`; the RSpectra
  generalized row includes the sparse Cholesky factorisation.
