---
title: "Benchmarks: where eigencore wins, loses, and crosses over"
output:
  rmarkdown::html_document:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
---

This is a pkgdown-only article, not a CRAN vignette: it runs `bench::mark()`
at build time and is excluded from `R CMD check` by the `^vignettes/articles$`
entry in `.Rbuildignore`.

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.6,
  fig.align = "center",
  out.width = "96%",
  dpi = 100
)

benchmark_ready <- all(vapply(
  c("bench", "RSpectra", "irlba"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))
benchmark_iterations <- 3L
options(knitr.kable.NA = "not run")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r helpers, include = FALSE}
`%||%` <- function(x, y) if (is.null(x)) y else x

fro_norm <- function(A) {
  if (inherits(A, "sparseMatrix")) {
    sqrt(sum(A@x^2))
  } else {
    sqrt(sum(abs(A)^2))
  }
}

relative_error <- function(x, truth) {
  x <- sort(Re(x), decreasing = TRUE)
  truth <- sort(Re(truth), decreasing = TRUE)
  max(abs(x - truth) / pmax(1, abs(truth)))
}

# Residual helpers keep sparse inputs sparse: densifying a 90 x 32000 operator
# just to measure a residual would dominate the measurement it reports on.
eigen_backward_error <- function(A, values, vectors) {
  values <- Re(values)
  vectors <- as.matrix(vectors)
  residual <- as.matrix(A %*% vectors) -
    vectors %*% diag(values, nrow = length(values))
  scale <- fro_norm(A) + abs(values)
  max(sqrt(colSums(abs(residual)^2)) / pmax(scale, .Machine$double.eps))
}

svd_backward_error <- function(A, d, u, v) {
  d <- Re(d)
  u <- as.matrix(u)
  v <- as.matrix(v)
  D <- diag(d, nrow = length(d))
  left_residual <- as.matrix(A %*% v) - u %*% D
  right_residual <- as.matrix(Matrix::crossprod(A, u)) - v %*% D
  residual <- sqrt(colSums(abs(left_residual)^2) +
                     colSums(abs(right_residual)^2))
  max(residual / pmax(fro_norm(A) + d, .Machine$double.eps))
}

bench_eval <- function(expr, iterations = benchmark_iterations) {
  expr <- substitute(expr)
  env <- parent.frame()
  last_result <- NULL
  mark <- tryCatch(
    bench::mark(
      last_result <- eval(expr, env),
      iterations = iterations,
      check = FALSE,
      memory = isTRUE(capabilities("profmem")),
      filter_gc = FALSE
    ),
    error = function(e) e
  )
  if (inherits(mark, "error")) {
    return(list(
      result = NULL,
      median_ms = NA_real_,
      mem_mb = NA_real_,
      error = conditionMessage(mark)
    ))
  }
  list(
    result = last_result,
    median_ms = as.numeric(mark$median[[1L]]) * 1000,
    mem_mb = as.numeric(mark$mem_alloc[[1L]]) / 1024^2,
    error = NA_character_
  )
}

make_dense_hermitian <- function(n, seed = 1L) {
  set.seed(seed)
  X <- matrix(rnorm(n * n), n, n)
  crossprod(X) / n + diag(seq(1, 1.2, length.out = n))
}

path_laplacian <- function(n) {
  Matrix::bandSparse(
    n,
    k = c(-1L, 0L, 1L),
    diagonals = list(rep(-1, n - 1L), c(1, rep(2, n - 2L), 1), rep(-1, n - 1L))
  )
}

random_sparse_hermitian <- function(n, density, seed = 1L) {
  set.seed(seed)
  A <- Matrix::rsparsematrix(n, n, density = density)
  methods::as(methods::as(A + Matrix::t(A), "generalMatrix"), "CsparseMatrix")
}

rectangular_sparse <- function(nrow, ncol, density = 0.035, seed = 1001L) {
  set.seed(seed)
  Matrix::rsparsematrix(nrow, ncol, density = density)
}

# RSpectra converges on the banded Laplacian only with a wider Krylov basis
# than its default. Giving it that basis keeps the comparison fair.
rspectra_eigen_opts <- function(tol, ncv = NULL) {
  opts <- list(tol = tol, maxitr = 10000L)
  if (!is.null(ncv)) opts$ncv <- ncv
  opts
}

eigen_rows <- function(name, A, k = 6L, tol = 1e-8, seed = 1L,
                       methods = c("eigencore", "RSpectra", "base"),
                       ncv = NULL, truth = NULL,
                       iterations = benchmark_iterations) {
  if (is.null(truth) && "base" %in% methods) {
    truth <- eigen(as.matrix(A), symmetric = TRUE,
                   only.values = TRUE)$values[seq_len(k)]
  }
  rows <- lapply(methods, function(method) {
    timed <- switch(
      method,
      eigencore = bench_eval({
        set.seed(seed)
        eig_partial(A, k = k, target = largest(), tol = tol)
      }, iterations = iterations),
      RSpectra = bench_eval({
        RSpectra::eigs_sym(A, k = k, which = "LA",
                           opts = rspectra_eigen_opts(tol, ncv))
      }, iterations = iterations),
      base = bench_eval({
        eigen(as.matrix(A), symmetric = TRUE)
      }, iterations = iterations)
    )
    if (!is.na(timed$error)) {
      return(data.frame(
        regime = name, task = "eigen", method = method,
        median_ms = timed$median_ms, mem_mb = timed$mem_mb,
        rel_error = NA_real_, backward_error = NA_real_,
        converged = FALSE, residual_check = FALSE,
        eigencore_label = NA_character_, status = timed$error,
        stringsAsFactors = FALSE
      ))
    }
    result <- timed$result
    extracted <- switch(
      method,
      eigencore = list(values = values(result), vectors = vectors(result),
                       nconv = result$nconv),
      RSpectra = list(values = result$values, vectors = result$vectors,
                      nconv = result$nconv),
      base = {
        ord <- order(result$values, decreasing = TRUE)[seq_len(k)]
        list(values = result$values[ord],
             vectors = result$vectors[, ord, drop = FALSE], nconv = k)
      }
    )
    converged <- isTRUE(extracted$nconv >= k)
    backward <- if (converged) {
      eigen_backward_error(A, extracted$values, extracted$vectors)
    } else {
      NA_real_
    }
    data.frame(
      regime = name, task = "eigen", method = method,
      median_ms = timed$median_ms, mem_mb = timed$mem_mb,
      rel_error = if (converged && !is.null(truth)) {
        relative_error(extracted$values, truth)
      } else {
        NA_real_
      },
      backward_error = backward,
      converged = converged,
      residual_check = isTRUE(backward <= tol),
      eigencore_label = if (method == "eigencore") result$method else NA_character_,
      status = if (converged) "ok" else "did not converge",
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

svd_rows <- function(name, A, rank = 6L, tol = 1e-8, seed = 1L,
                     methods = c("eigencore", "RSpectra", "irlba", "base"),
                     truth = NULL, iterations = benchmark_iterations) {
  if (is.null(truth) && "base" %in% methods) {
    truth <- svd(as.matrix(A), nu = 0, nv = 0)$d[seq_len(rank)]
  }
  rows <- lapply(methods, function(method) {
    timed <- switch(
      method,
      eigencore = bench_eval({
        set.seed(seed)
        svd_partial(A, rank = rank, target = largest(), tol = tol)
      }, iterations = iterations),
      RSpectra = bench_eval({
        RSpectra::svds(A, k = rank, nu = rank, nv = rank,
                       opts = list(tol = tol, maxitr = 10000L))
      }, iterations = iterations),
      irlba = bench_eval({
        set.seed(seed)
        irlba::irlba(A, nv = rank, nu = rank, tol = tol)
      }, iterations = iterations),
      base = bench_eval({
        svd(as.matrix(A), nu = rank, nv = rank)
      }, iterations = iterations)
    )
    if (!is.na(timed$error)) {
      return(data.frame(
        regime = name, task = "SVD", method = method,
        median_ms = timed$median_ms, mem_mb = timed$mem_mb,
        rel_error = NA_real_, backward_error = NA_real_,
        converged = FALSE, residual_check = FALSE,
        eigencore_label = NA_character_, status = timed$error,
        stringsAsFactors = FALSE
      ))
    }
    result <- timed$result
    extracted <- switch(
      method,
      eigencore = list(d = values(result), u = left_vectors(result),
                       v = right_vectors(result)),
      RSpectra = list(d = result$d, u = result$u, v = result$v),
      irlba = list(d = result$d, u = result$u, v = result$v),
      base = list(d = result$d[seq_len(rank)], u = result$u, v = result$v)
    )
    converged <- length(extracted$d) >= rank &&
      all(is.finite(extracted$d[seq_len(rank)]))
    backward <- if (converged) {
      svd_backward_error(A, extracted$d, extracted$u, extracted$v)
    } else {
      NA_real_
    }
    data.frame(
      regime = name, task = "SVD", method = method,
      median_ms = timed$median_ms, mem_mb = timed$mem_mb,
      rel_error = if (converged && !is.null(truth)) {
        relative_error(extracted$d, truth)
      } else {
        NA_real_
      },
      backward_error = backward,
      converged = converged,
      residual_check = isTRUE(backward <= tol),
      eigencore_label = if (method == "eigencore") result$method else NA_character_,
      status = if (converged) "ok" else "did not converge",
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

metric_table <- function(rows, metric, digits = 4L,
                         methods = c("eigencore", "RSpectra", "irlba", "base")) {
  regimes <- unique(rows$regime)
  out <- data.frame(regime = regimes, check.names = FALSE)
  for (method in methods) {
    out[[method]] <- vapply(regimes, function(regime) {
      keep <- rows$regime == regime & rows$method == method &
        rows$status == "ok"
      if (!any(keep)) NA_real_ else rows[[metric]][which(keep)[1L]]
    }, numeric(1))
    out[[method]] <- signif(out[[method]], digits)
  }
  out
}

speedup_column <- function(rows, reference = "RSpectra") {
  regimes <- unique(rows$regime)
  vapply(regimes, function(regime) {
    ec <- rows$median_ms[rows$regime == regime &
                           rows$method == "eigencore" & rows$status == "ok"]
    rf <- rows$median_ms[rows$regime == regime &
                           rows$method == reference & rows$status == "ok"]
    if (!length(ec) || !length(rf)) return(NA_character_)
    ratio <- rf[[1L]] / ec[[1L]]
    if (ratio >= 1) {
      sprintf("%.1fx faster", ratio)
    } else {
      sprintf("%.1fx slower", 1 / ratio)
    }
  }, character(1))
}

quality_table <- function(rows) {
  regimes <- unique(rows$regime)
  do.call(rbind, lapply(regimes, function(regime) {
    current <- rows[rows$regime == regime, , drop = FALSE]
    ok <- current[current$status == "ok", , drop = FALSE]
    if (!nrow(ok)) {
      return(data.frame(
        regime = regime,
        successful_methods = sprintf("0/%d", nrow(current)),
        max_relative_error = NA_character_,
        max_backward_error = NA_character_,
        residual_checks = "not available",
        stringsAsFactors = FALSE
      ))
    }
    rel <- ok$rel_error[is.finite(ok$rel_error)]
    data.frame(
      regime = regime,
      successful_methods = sprintf("%d/%d", nrow(ok), nrow(current)),
      max_relative_error = if (length(rel)) {
        formatC(max(rel), format = "e", digits = 2)
      } else {
        "not compared"
      },
      max_backward_error = formatC(max(ok$backward_error), format = "e",
                                   digits = 2),
      residual_checks = if (all(ok$residual_check)) "all pass" else "one or more fail",
      stringsAsFactors = FALSE
    )
  }))
}
```

Whether eigencore is faster than `RSpectra` or `irlba` depends almost entirely
on the problem, and the dependence is strong enough that a single headline
number would be misleading. This vignette measures three regimes that land in
three different places, then varies one dimension at a time to locate the
crossover between them.

Everything below describes one render on one machine. Treat the ratios as
illustrations of shape, not as constants.

```{r setup}
library(eigencore)
```

## Three regimes, three outcomes

eigencore is not a faster implementation of the same algorithm that `RSpectra`
runs. It inspects the operator, and when it recognizes exploitable structure it
solves a different, cheaper problem. Where no such structure exists, it runs a
thick-restart Lanczos iteration that has no particular advantage over ARPACK,
which is what `RSpectra` wraps.

That produces three regimes:

```{r regimes}
benchmark_regimes <- data.frame(
  regime = c(
    "banded Hermitian",
    "wide sparse SVD",
    "dense Hermitian",
    "general sparse Hermitian"
  ),
  input = c(
    "1500 x 1500 tridiagonal",
    "90 x 50000 dgCMatrix",
    "1000 x 1000 dense",
    "2000 x 2000 dgCMatrix, 1% dense"
  ),
  structure_used = c(
    "banded factorization, shift-invert",
    "bounded Gram problem",
    "none",
    "none"
  )
)
knitr::kable(benchmark_regimes)
```

```{r build-headline, eval = benchmark_ready, include = FALSE}
headline_cases <- list(
  banded = path_laplacian(1500L),
  wide = rectangular_sparse(90L, 50000L),
  dense = make_dense_hermitian(1000L, seed = 1001L),
  general = random_sparse_hermitian(2000L, density = 0.01, seed = 7L)
)
```

```{r run-headline, eval = benchmark_ready, include = FALSE}
headline_rows <- rbind(
  eigen_rows("banded Hermitian", headline_cases$banded,
             k = 6L, seed = 2001L, ncv = 60L),
  svd_rows("wide sparse SVD", headline_cases$wide,
           rank = 6L, seed = 2002L),
  eigen_rows("dense Hermitian", headline_cases$dense,
             k = 6L, seed = 2003L),
  eigen_rows("general sparse Hermitian", headline_cases$general,
             k = 6L, seed = 2004L,
             methods = c("eigencore", "RSpectra"),
             truth = eigen(as.matrix(headline_cases$general), symmetric = TRUE,
                           only.values = TRUE)$values[seq_len(6L)])
)
```

```{r headline-table, eval = benchmark_ready, echo = FALSE}
headline_timing <- metric_table(headline_rows, "median_ms")
headline_timing$vs_rspectra <- speedup_column(headline_rows)
knitr::kable(
  headline_timing[, c("regime", "vs_rspectra", "eigencore", "RSpectra",
                      "irlba", "base")],
  col.names = c("case", "eigencore vs RSpectra", "eigencore", "RSpectra",
                "irlba", "base R"),
  align = c("l", "l", "r", "r", "r", "r"),
  caption = paste("Median solver-call time in milliseconds from",
                  benchmark_iterations,
                  "iterations. The comparison column is RSpectra time divided",
                  "by eigencore time.")
)
```

The two wins come from structure, not from a faster inner loop. The banded case
is the clearest. The largest eigenvalues of a path Laplacian are separated by
gaps on the order of `1e-5`, which is punishing for a method that sees only a
matrix-vector product: ARPACK has to grind through the cluster. Eigencore
instead detects a tridiagonal operator, factorizes it, and runs shift-invert
Lanczos, which converges in a handful of iterations regardless of how tight the
cluster is.

Because that cluster is what makes the case hard, the comparison is run with
`ncv = 60` rather than the `RSpectra` default. With defaults `RSpectra` still
converges at this size, but it needs roughly ten times as many iterations and is
about 60% slower, so the wider basis is the more favorable setting for it. At
`n = 2000` and above, the default basis stops converging within `maxitr`
entirely.

The two losses are equally structural. On a dense matrix and on a general
sparse matrix there is nothing to exploit, and eigencore's Lanczos
implementation is slower than ARPACK's per iteration.

## Where the crossover is

The rectangular sparse SVD case is the one worth studying, because it has a
clean crossover. eigencore forms the smaller Gram problem and certifies the
resulting triplets back in the original coordinates. If the small side stays
bounded, that Gram problem does not grow as the long side does, while an
iterative method pays for every matrix-vector product against a matrix that
keeps getting longer.

The sweep below fixes everything except the long dimension: small side 90, rank
6, density 0.035, same tolerance, same fixture generator.

```{r crossover, eval = benchmark_ready, include = FALSE}
crossover_long_sides <- c(200L, 500L, 2000L, 8000L, 32000L)

# base R is excluded from the timed methods here because timing a dense
# decomposition of a 90 x 32000 operator says nothing useful. It is still used
# once per case, outside the timer, to supply reference singular values.
dense_truth <- function(A, rank) svd(as.matrix(A), nu = 0, nv = 0)$d[seq_len(rank)]

crossover_rows <- do.call(rbind, lapply(crossover_long_sides, function(n) {
  wide <- rectangular_sparse(90L, n)
  tall <- rectangular_sparse(n, 90L)
  rbind(
    svd_rows(sprintf("wide 90 x %d", n), wide, rank = 6L, seed = 3001L,
             methods = c("eigencore", "RSpectra", "irlba"),
             truth = dense_truth(wide, 6L)),
    svd_rows(sprintf("tall %d x 90", n), tall, rank = 6L, seed = 3002L,
             methods = c("eigencore", "RSpectra", "irlba"),
             truth = dense_truth(tall, 6L))
  )
}))
```

```{r crossover-table, eval = benchmark_ready, echo = FALSE}
crossover_timing <- metric_table(
  crossover_rows, "median_ms",
  methods = c("eigencore", "RSpectra", "irlba")
)
crossover_timing$vs_rspectra <- speedup_column(crossover_rows)
crossover_timing <- crossover_timing[
  order(!grepl("^wide", crossover_timing$regime)),
]
knitr::kable(
  crossover_timing[, c("regime", "vs_rspectra", "eigencore", "RSpectra",
                       "irlba")],
  row.names = FALSE,
  col.names = c("case", "eigencore vs RSpectra", "eigencore", "RSpectra",
                "irlba"),
  align = c("l", "l", "r", "r", "r"),
  caption = "Median solver-call time in milliseconds, small side fixed at 90."
)
```

```{r crossover-summary, eval = benchmark_ready, echo = FALSE, results = 'asis'}
crossing <- function(prefix) {
  sub <- crossover_rows[grepl(prefix, crossover_rows$regime), , drop = FALSE]
  sides <- crossover_long_sides
  won <- vapply(sides, function(n) {
    regime <- unique(sub$regime[grepl(sprintf("\\b%d\\b", n), sub$regime)])
    if (!length(regime)) return(NA)
    ec <- sub$median_ms[sub$regime == regime[[1L]] & sub$method == "eigencore"]
    rs <- sub$median_ms[sub$regime == regime[[1L]] & sub$method == "RSpectra"]
    if (!length(ec) || !length(rs)) return(NA)
    ec[[1L]] < rs[[1L]]
  }, logical(1))
  first <- which(won)[1L]
  if (is.na(first)) "did not cross in this sweep" else
    sprintf("between %d and %d", sides[max(first - 1L, 1L)], sides[first])
}
cat(sprintf(
  paste0(
    "In this render the wide orientation crossed **%s**, and the tall ",
    "orientation crossed **%s**.\n"
  ),
  crossing("^wide"),
  crossing("^tall")
))
```

Both orientations contract the long side into a bounded Gram problem of the same
size, and both cross.

Where they cross is not a property of eigencore alone, so do not carry the
number above to a different matrix. It moves with density: this sweep uses
0.035, while the repository's controlled sweep in
`inst/benchmarks/bench-svd-gram-cutoff.R` holds the small side at 90 as well but
uses 0.003, about twelve times sparser, and there the tall orientation does not
cross on raw solver time until a long side near 25,000. Fewer nonzeros per
column make an iterative matrix-vector product cheaper while eigencore's Gram
formation is comparatively less affected, which pushes the crossover out.

It also moves with eigencore's own per-call fixed costs, which is the subject of
the next section and is the larger effect at the time of writing.

Below the crossover, eigencore's fixed costs are simply larger than the whole
problem. That is not an artifact of the measurement, and it does not go away
with tuning.

## What the fixed costs are

A `svd_partial()` or `eig_partial()` call does work that `RSpectra::svds()` does
not do at all:

- **It checks the operator instead of trusting a claim.** `RSpectra::eigs_sym()`
  is symmetric because of its name; you assert the property by choosing the
  function. `eig_partial()` verifies it, which for a dense `n x n` input is an
  `O(n^2)` scan.
- **It plans.** The planner inspects structure, target, and requested rank to
  choose a kernel, and records the choice.
- **It records provenance.** Operators carry an identity and revision so that
  results, restart states, and factorizations can be traced to the operator that
  produced them. Deriving that identity for a built-in operator digests the
  source data, so it also scales with the size of the input rather than with the
  amount of solving requested.
- **It certifies.** Every result carries residuals, a backward-error bound, and
  an orthogonality measure. The comparison methods return none of these, and the
  residual checks used elsewhere in this article run outside the timer.

Only the last of those is about the answer. The rest is bookkeeping that a
single call pays in full, which is why the small end of every sweep above goes
to the competition.

The breakdown below separates them on the dense case. `kernel` is the solver's
own instrumented stage total, so the gap between it and the full call is
everything else.

```{r fixed-cost-breakdown, eval = benchmark_ready, echo = FALSE}
dense_A <- headline_cases$dense
dense_op <- as_operator(dense_A)
fixed_cost <- bench::mark(
  full = {
    set.seed(11)
    eig_partial(dense_A, k = 6L, target = largest(), tol = 1e-8)
  },
  prebuilt = {
    set.seed(11)
    eig_partial(dense_op, k = 6L, target = largest(), tol = 1e-8)
  },
  uncertified = {
    set.seed(11)
    eig_partial(dense_A, k = 6L, target = largest(), tol = 1e-8, certify = FALSE)
  },
  iterations = benchmark_iterations,
  check = FALSE,
  memory = FALSE,
  filter_gc = FALSE
)
set.seed(11)
dense_fit <- eig_partial(dense_A, k = 6L, target = largest(), tol = 1e-8)
knitr::kable(
  data.frame(
    measurement = c(
      "full call, matrix input",
      "call with a prebuilt operator",
      "call with certify = FALSE",
      "solver kernel stages only"
    ),
    median_ms = signif(c(
      as.numeric(fixed_cost$median) * 1000,
      sum(dense_fit$stage_seconds) * 1000
    ), 4)
  ),
  col.names = c("measurement", "median ms"),
  align = c("l", "r"),
  caption = paste("Same", nrow(dense_A), "x", ncol(dense_A),
                  "dense Hermitian problem, k = 6.")
)
```

Two things follow. Certification is not the expensive part: turning it off
changes almost nothing. And building the operator once with `as_operator()` and
reusing it removes a repeated share of the bookkeeping, which is worth doing
whenever you solve more than one problem against the same matrix.


## Allocated memory

These values are megabytes allocated during the timed expression. They do not
measure peak resident memory.

```{r memory-table, eval = benchmark_ready, echo = FALSE}
knitr::kable(
  metric_table(rbind(headline_rows, crossover_rows), "mem_mb"),
  col.names = c("case", "eigencore", "RSpectra", "irlba", "base R"),
  align = c("l", "r", "r", "r", "r"),
  caption = "Allocated memory in megabytes."
)
```

## Numerical checks

Timing means nothing without agreement on the answer. The summary below reports
the largest error observed across the methods in each case. Relative error is
compared against a dense `base` reference where computing one is affordable; the
backward error is always measured directly on the returned vectors.

```{r quality-table, eval = benchmark_ready, echo = FALSE}
knitr::kable(
  quality_table(rbind(headline_rows, crossover_rows)),
  row.names = FALSE,
  col.names = c("case", "methods completed", "max relative error",
                "max backward error", "residual checks"),
  align = c("l", "c", "r", "r", "l"),
  caption = "Numerical checks across all methods in each case."
)
```

## Eigencore planner paths

The planner label records which kernel produced each result, so a timing change
can be traced to a dispatch change.

```{r planner-table, eval = benchmark_ready, echo = FALSE}
planner_rows <- rbind(headline_rows, crossover_rows)
planner_rows <- planner_rows[
  planner_rows$method == "eigencore",
  c("regime", "eigencore_label", "status"),
  drop = FALSE
]
rownames(planner_rows) <- NULL
knitr::kable(
  planner_rows,
  col.names = c("case", "planner label", "status"),
  align = c("l", "l", "l")
)
```

## How to read this

The honest summary is that eigencore is worth reaching for when one of these is
true:

- the operator has structure it can exploit (banded, tridiagonal, diagonal,
  separable grid, or a bounded Gram side on a highly rectangular sparse matrix);
- you want a certificate with the answer, and would otherwise write the
  residual check yourself;
- the spectrum is clustered enough that a generic Krylov method needs tuning to
  converge at all, as `RSpectra` does on the banded case above once `n` passes
  about 2000.

For a dense symmetric matrix, or a general sparse matrix with no exploitable
structure, `RSpectra` is faster and this vignette should not be read as
suggesting otherwise.

## Limits of this comparison

These cases are small enough to rebuild on every render. They catch obvious
problems: wrong target, broken sparse handling, missing vectors, certification
drift, and large timing regressions. They do not establish scaling behavior, and
a single crossover measured on one machine is not a constant.

Two caveats matter for interpreting any row:

- `base` R is a truth oracle only while `as.matrix(A)` is affordable, which is
  why it is absent from the larger crossover rows.
- A single-vector Krylov method may return fewer copies of a repeated eigenvalue
  than its true multiplicity. Eigencore's certificate attests that each returned
  pair is a genuine eigenpair; it does not attest that no eigenvalue was missed.
  Benchmark fixtures here avoid degenerate spectra for that reason.

The benchmark tables report `bench::mem_alloc`, which counts cumulative
R-managed allocation. It does not see C/C++ working heaps used by eigencore or
external native solvers, so it must not be read as cross-engine peak memory.
The installed Hermitian G1 gate uses a narrower, explicit memory claim: the
retained solver result may be at most 1.25 times the result from the smallest
certified reference, allowing bounded space for eigencore's plan, certificate,
and provenance metadata. Cumulative R allocation remains visible as a
diagnostic; a total or peak-memory claim needs a separate process-level RSS
measurement.

## Run the larger benchmark scripts

The scripts under `inst/benchmarks` run larger cases with stricter checks and
saved outputs. Run them from an installed package rather than a mutable source
tree:

```sh
R CMD INSTALL --library=/tmp/eigencore-bench-lib .
R_LIBS=/tmp/eigencore-bench-lib \
  Rscript inst/benchmarks/bench-native-hermitian-gate.R \
  --strict --save --cases=path_laplacian:1000
```

For the SVD surface, compare eigencore against `RSpectra`, `irlba`, and base
rows where those baselines are meaningful:

```sh
R_LIBS=/tmp/eigencore-bench-lib \
  Rscript inst/benchmarks/bench-svd-surface.R \
  --iterations=1 --h-candidate \
  --methods=eigencore,RSpectra,irlba,base \
  --cases=tall_sparse,wide_sparse \
  --subject=eigencore --strict --save
```

The controlled mode of `bench-svd-gram-cutoff.R` fixes the small side, rank,
density, and fixture while varying only the long side, and reports raw solver
and common certified-total ratios separately. Use
`docs/v1-benchmark-manifest.md` for the current release-gate inventory and
`docs/post-v1-benchmark-gates.md` for future-promotion gates.
